\section{Experimental Design}
\label{sec:experiments}

This section describes the experimental protocols, baseline schemes, training procedures, and hardware parameters used to validate the proposed hybrid quantization-TCN framework. All experiments are conducted on the 10-site DALTON subset described in Section~\ref{sec:dataset}, using chronological train-validation-test splits to ensure temporal integrity.

\subsection{Baseline Schemes}

To isolate the contributions of sensor-side quantization, temporal subsampling, and TCN architecture, we define six baseline and proposed schemes for comparative evaluation:

\subsubsection{Baseline 1: Full-Precision Baseline (FP-Baseline)}
Transmits 16-bit quantized sensor readings for all six pollutant channels at 1 Hz ($T_S = 1$ s), yielding 96 bits per packet. The TCN model is trained on high-precision data (quantization noise negligible compared to sensor intrinsic noise). This scheme represents conventional IoT monitoring without compression and establishes the upper bound on prediction accuracy.

\subsubsection{Baseline 2: Uniform 8-bit Quantization (UQ-8)}
Applies static 8-bit uniform quantization per channel with ranges $[a_c, b_c]$ set to the global minimum and maximum observed across all training data. Transmission occurs at 1 Hz, yielding 48 bits per packet. This baseline evaluates naive bit-width reduction without adaptive range scaling or temporal subsampling, representing common practice in resource-constrained IoT deployments \cite{taştan2022lowcost}.

\subsubsection{Baseline 3: Downsampling with Averaging (DS-4s)}
Transmits 16-bit quantized readings every 4 seconds, where each transmitted value is the arithmetic mean of the preceding 4 samples. This reduces transmission frequency by 4$\times$ without reducing per-sample precision, yielding 96 bits per packet but average transmission energy of 3.2 $\mu$J/s (see Section~\ref{sec:methodology}). The TCN receives downsampled inputs at effective 0.25 Hz sampling rate.

\subsubsection{Proposed Scheme 1: Hybrid Q6$\Delta$ (H-Q6$\Delta$)}
Implements 6-bit residual quantization (Equation~\eqref{eq:residual_quantization}) with adaptive range scaling (Equation~\eqref{eq:adaptive_range}) updated every $T_R = 300$ s. Transmission occurs every $T_S = 4$ s, yielding 36 bits per packet. Absolute reference values are transmitted every $N_{\text{ref}} = 15$ packets (60 s) to prevent drift, adding 96 bits every 60 s (amortized overhead: 1.6 bits/packet). The TCN is trained using quantization-aware training (QAT) with simulated 6-bit residual noise.

\subsubsection{Proposed Scheme 2: Hybrid Q4$\Delta$ (H-Q4$\Delta$)}
Identical to H-Q6$\Delta$ but uses 4-bit residual quantization, yielding 24 bits per packet. This scheme represents the most aggressive quantization configuration, prioritizing energy savings while accepting potential accuracy degradation. The residual dynamic range is set to $[-4\sigma_c, +4\sigma_c]$ to accommodate 99.99\% of residuals, preventing clipping for all but extreme transient events.

\subsubsection{Proposed Scheme 3: Hybrid Mixed-Precision (H-Mix)}
Allocates heterogeneous bit-widths across channels based on their predictive importance for PM$_{2.5}$ forecasting, determined via ablation studies. Specifically: PM$_{10}$, CO, and VOC (high importance) receive 6 bits each, while CO$_2$, NO$_2$, and PM$_{2.5}$ itself (moderate importance) receive 4 bits each, yielding 30 bits per packet. This mixed-precision strategy balances energy efficiency and accuracy by prioritizing informative channels.

\subsection{Training Protocols}

\subsubsection{Model Architecture and Hyperparameters}

The TCN architecture specified in Table~\ref{tab:tcn_architecture} is implemented in TensorFlow 2.15 using custom dilated causal convolutional layers. Hyperparameters are fixed across all schemes to ensure fair comparison:

\begin{itemize}
\item \textbf{Optimizer}: Adam with learning rate $\alpha = 10^{-3}$, exponential decay schedule (decay rate 0.95 every 10 epochs).
\item \textbf{Batch size}: 64 windows per batch, shuffled within each epoch while preserving site identity.
\item \textbf{Regularization}: Spatial dropout rate 0.2 applied after each convolutional layer; L2 weight decay $\lambda = 10^{-5}$.
\item \textbf{Loss function}: Mean squared error (MSE) on normalized targets.
\item \textbf{Early stopping}: Monitor validation RMSE with patience of 15 epochs; restore weights from best validation epoch.
\item \textbf{Training duration}: Maximum 100 epochs, typically converges within 40-60 epochs.
\end{itemize}

For quantization-aware training (QAT), quantization operators $Q_c(\cdot)$ are inserted into the input pipeline, simulating the effect of reduced precision during forward passes. Gradients flow through straight-through estimators (STE) as described in Equation~\eqref{eq:ste}. Each quantization configuration (4-bit, 6-bit, mixed) trains a separate model instance, allowing specialization to the specific noise characteristics.

\subsubsection{Data Augmentation}

To improve robustness to sensor drift and calibration errors, we apply mild Gaussian noise augmentation during training:
\begin{equation}
\tilde{x}_t^{(c)} \leftarrow \tilde{x}_t^{(c)} + \epsilon, \quad \epsilon \sim \mathcal{N}(0, 0.05\sigma_c),
\label{eq:augmentation}
\end{equation}
where $\sigma_c$ is the per-channel standard deviation. This augmentation simulates quantization noise beyond that introduced by the quantizer itself, enhancing generalization to unseen sites and conditions \cite{kim2025machine}.

\subsubsection{Post-Training Quantization of TCN Weights}

After training in FP32, each TCN model is converted to INT8 using TensorFlow Lite's dynamic range quantization. A representative calibration dataset comprising 1,000 randomly sampled validation windows is used to determine per-layer scale factors and zero-points. The quantized TFLite model is validated on the same validation set to confirm $<$0.05 $\mu$g/m$^3$ RMSE degradation relative to FP32 inference, consistent with prior findings \cite{mazinani2025air}.

\subsection{Hardware Parameters and Energy Model}

\subsubsection{Wireless Communication Parameters}

Energy calculations in Equation~\eqref{eq:energy_per_second} use representative values for low-power wireless transceivers commonly deployed in IoT air quality systems:

\begin{itemize}
\item \textbf{Radio energy per bit}: $E_{\text{tx}} = 50$ nJ/bit, corresponding to nRF52840 BLE transceiver at $-$4 dBm transmit power \cite{khan2024artificial}.
\item \textbf{Protocol overhead}: $b_{\text{overhead}} = 160$ bits (20-byte MAC/PHY headers for BLE advertising packets).
\item \textbf{Packet transmission time}: $\approx$ 1-2 ms at 1 Mbps BLE data rate, negligible compared to $T_S = 4$ s intervals.
\end{itemize}

For schemes with adaptive range scaling, metadata packets containing updated $\{a_c(t), b_c(t)\}$ values (24 bytes = 192 bits for 6 channels with 32-bit floats) are transmitted every $T_R = 300$ s, adding 0.64 bits/s amortized overhead.

\subsubsection{Sensing and Computation Energy}

Sensing energy $E_{\text{sense}}^{(c)}$ for gas sensors (MOS, electrochemical) is dominated by heater power and ranges from 5-20 mW depending on sensor type \cite{przystupa2025ensuring}. Assuming continuous sensing (no duty cycling), per-second sensing energy is constant across schemes:
\begin{equation}
E_{\text{sense,total}} = \sum_{c=1}^{6} E_{\text{sense}}^{(c)} \approx 15 \text{ mW} \cdot 1 \text{ s} = 15 \text{ mJ/s}.
\label{eq:sensing_energy}
\end{equation}

Quantization computation on a 32-bit ARM Cortex-M4 MCU (e.g., STM32L476) requires $\approx$ 2,000 cycles per packet (integer arithmetic, percentile updates amortized over $T_R$), consuming:
\begin{equation}
E_{\text{MCU}} = 2{,}000 \text{ cycles} \times 0.1 \text{ nJ/cycle} / 4 \text{ s} \approx 0.05 \text{ } \mu\text{J/s}.
\label{eq:mcu_energy}
\end{equation}

Since $E_{\text{MCU}} \ll E_{\text{sense}}$ and $E_{\text{MCU}} \ll E_{\text{tx}}$, we focus comparative analysis on transmission energy, which exhibits the largest variance across schemes (Table~\ref{tab:energy_comparison}).

\subsubsection{Edge Gateway Inference}

TCN inference on a Raspberry Pi 4 (quad-core ARM Cortex-A72, 4 GB RAM) using TensorFlow Lite INT8 models achieves 8-12 ms latency per window, consuming $\approx$ 150 mW during inference bursts. Averaged over $T_S = 4$ s intervals and multiple concurrent sensor nodes (e.g., 10 nodes), per-node inference energy is $\approx$ 0.4 mJ/s, an order of magnitude lower than sensor-side transmission costs. Therefore, our energy analysis focuses on sensor-to-gateway transmission, which dominates end-to-end system energy budgets.

\subsection{Evaluation Metrics}

\subsubsection{Prediction Accuracy}

Primary metrics for forecasting performance are:
\begin{itemize}
\item \textbf{Root Mean Squared Error (RMSE)}: Penalizes large prediction errors, critical for detecting hazardous pollution events.
\item \textbf{Mean Absolute Error (MAE)}: Provides interpretable average deviation in $\mu$g/m$^3$.
\item \textbf{R$^2$ Score}: Measures fraction of variance explained by the model, indicating goodness-of-fit.
\end{itemize}

Metrics are computed per site and averaged across the 10-site test set to account for inter-site variability.

\subsubsection{Energy Efficiency}

Energy efficiency is quantified via:
\begin{itemize}
\item \textbf{Average energy per second} ($E_{\text{sec}}$, $\mu$J/s): Combines sensing, computation, and communication per Equation~\eqref{eq:energy_per_second}, with sensing energy treated as constant.
\item \textbf{Energy reduction} (\%): Relative savings compared to FP-Baseline.
\item \textbf{Energy-accuracy product}: $E_{\text{sec}} \times \text{RMSE}$, a composite metric balancing energy and accuracy (lower is better).
\end{itemize}

\subsubsection{Robustness and Generalizability}

To assess generalizability, we report:
\begin{itemize}
\item \textbf{Per-site RMSE variance}: Standard deviation of RMSE across 10 test sites, indicating sensitivity to site-specific characteristics.
\item \textbf{Seasonal performance}: Separate RMSE for winter (October-December) and summer (February-March) test periods at sites with sufficient seasonal coverage.
\item \textbf{Activity-conditioned accuracy}: RMSE stratified by DALTON activity labels (cooking, cleaning, quiescent periods) where annotations are available.
\end{itemize}

\subsection{Ablation Studies}

To isolate the contributions of individual components, we conduct ablation experiments:

\begin{enumerate}
\item \textbf{Adaptive vs. Static Range Scaling}: Compare H-Q6$\Delta$ with adaptive range updates (Equation~\eqref{eq:adaptive_range}) versus fixed ranges computed over entire training set.

\item \textbf{Residual vs. Absolute Quantization}: Compare H-Q6$\Delta$ (residual encoding per Equation~\eqref{eq:residual_quantization}) versus a variant transmitting absolute quantized values at the same 6-bit precision and $T_S = 4$ s interval.

\item \textbf{Quantization-Aware vs. Post-Hoc Training}: Compare H-Q6$\Delta$ with QAT (quantization operators in training loop) versus training on full-precision data and applying quantization only during inference.

\item \textbf{TCN vs. LSTM Baseline}: Replace the TCN with a 2-layer LSTM (64 hidden units per layer, dropout 0.3) and evaluate accuracy and inference latency on Raspberry Pi 4.

\item \textbf{Mixed-Precision Bit Allocation}: Evaluate H-Mix variants with different channel-wise bit-width assignments (e.g., uniform 5-bit, prioritizing different channels) to validate the chosen allocation.
\end{enumerate}

\subsection{Implementation Details}

\subsubsection{Software Environment}
\begin{itemize}
\item \textbf{Training}: TensorFlow 2.15, Python 3.10, conducted on an NVIDIA RTX 4090 GPU (24 GB VRAM).
\item \textbf{Inference}: TensorFlow Lite 2.15 on Raspberry Pi 4 Model B (Raspbian OS 11, ARM Cortex-A72).
\item \textbf{Data processing}: NumPy 1.24, Pandas 2.0 for preprocessing and metric computation.
\end{itemize}

\subsubsection{Reproducibility}
All code, trained models, and preprocessed DALTON subsets are available in the project repository at \url{https://github.com/drhamzahfaraj/sensor-quantization-air-quality}. Random seeds are fixed (seed = 42) for data splitting, weight initialization, and training shuffling to ensure reproducibility.

\subsection{Summary}

This experimental design enables rigorous evaluation of the proposed hybrid quantization-TCN framework across multiple dimensions: prediction accuracy, energy efficiency, robustness, and generalizability. The next section presents comprehensive results and ablation analyses.