\begin{abstract}
Indoor air pollution contributes to approximately 3.2 million premature deaths annually, necessitating continuous monitoring through low-cost Internet of Things (IoT) sensor networks. However, resource-constrained edge devices face a fundamental tension between data fidelity and energy sustainability. This paper presents Variance-Aware Adaptive Bit-Width Quantisation (VABQ), a unified framework that jointly optimises sensor-level analogue-to-digital converter (ADC) resolution and inference-level neural network precision for energy-efficient air quality monitoring. VABQ operates in two stages: Stage~1 adapts per-channel ADC bit-width (4, 8, or 16~bits) based on a variance-entropy criterion that exploits temporal signal structure, refined by wireless channel state information to prevent over-quantisation during poor link conditions; Stage~2 applies INT8 post-training quantisation to a lightweight one-dimensional convolutional neural network for pollutant event classification. We introduce intra-window sample-adaptive precision allocation that detects transient spikes (cooking, ventilation events) and temporarily boosts bit-width by 4~bits, improving event detection F1-score from 94.1\% to 97.8\%. Daily Bayesian optimisation with Gaussian process surrogates adapts hyperparameters to seasonal drift, maintaining $96.1 \pm 0.4$\% accuracy over six-month deployments versus 94.7\% degradation with static thresholds. Evaluated on the DALTON dataset---30 indoor sites, 89.1 million samples spanning six months across summer and winter seasons in developing-country contexts---VABQ achieves 64.5\% ADC energy reduction, 74.8\% inference energy reduction, and 16.7\% transmission time reduction. However, microcontroller idle power dominates total system energy (10.44~J per 10-minute window versus 2.30~J inference baseline), yielding 13.5\% total system energy savings (12.746~J $\rightarrow$ 11.025~J) while maintaining 97.4\% classification accuracy and 2.1\% normalised root-mean-square reconstruction error. Comparative analysis against ten state-of-the-art baselines confirms VABQ achieves Pareto optimality---highest accuracy at lowest energy---with statistical significance ($p < 0.0001$, paired $t$-test). The framework's principal novelty lies in bridging traditionally separate sensor compression and neural network quantisation domains through channel-state-aware multi-scale precision control, enabling compound energy savings that exceed independent optimisation of sensing, transmission, and inference layers.
\end{abstract}

\noindent\textbf{Keywords:} Edge AI; dynamic quantisation; air quality monitoring; Internet of Things; post-training quantisation; adaptive sampling; Bayesian optimisation; LoRaWAN; temporal convolutional networks; hardware-aware neural architecture

\newpage
