\section{Experimental Setup}
\label{sec:experiments}

\subsection{The DALTON Dataset}
\label{sec:exp_dataset}

Experiments use the DALTON dataset collected over six months across 30 indoor deployment sites in four geographical regions of India~\cite{karmakar2024exploring, karmakar2024indoor}. Sites comprise residential households (39.3\%), studio apartments (28.6\%), research laboratories (17.9\%), food canteens (7.1\%), and classrooms (7.1\%). Site areas range from 150 to 1,100~sq~ft, with 1--6 sensors per site and 46 occupants (42 actively annotating).

The DALTON sensor node is built on the ESP-WROOM-32 chip (dual-core Xtensa LX6, 80--240~MHz, 448~KB ROM, 520~KB SRAM). It measures eight pollutant-related channels at 1~Hz:
\begin{itemize}[leftmargin=*]
    \item \textbf{PM$_1$/PM$_{2.5}$/PM$_{10}$}: DUST sensor, range 0--500~$\mu$g/m$^3$, error $\pm 10$~$\mu$g/m$^3$.
    \item \textbf{CO$_2$}: MH-Z16 NDIR sensor, range 0--10,000~ppm, error $\pm 100$~ppm + 6\%.
    \item \textbf{NO$_2$, VOC, CO, C$_2$H$_5$OH}: Multi-Channel Gas Sensor (MCGS), ranges 0.1--10~ppm (NO$_2$), 1--500~ppm (ethanol), 5--5,000~ppm (CO).
    \item \textbf{Temperature, Humidity}: DHT22, $\pm 0.5$°C, $\pm 2$\% RH.
\end{itemize}
The node draws a maximum of 3.55~W (760~mA at 5~V DC), measures $112 \times 112 \times 55$~mm, weighs 160~g, and costs approximately \$250~\cite{karmakar2024exploring}. WiFi 2.4~GHz provides connectivity.

\subsection{Data Partitioning}
\label{sec:exp_partition}

Data is split temporally to preserve non-stationarity and prevent future leakage:
\begin{itemize}[leftmargin=*]
    \item \textbf{Training (Months 1--3):} Model training and VABQ calibration ($\theta_L$, $\theta_H$, $\alpha$). The first week serves as calibration subset for computing $\sigma_{c,\max}^2$ and $H_{c,\max}$.
    \item \textbf{Validation (Month 4):} Hyperparameter tuning and early stopping.
    \item \textbf{Testing (Months 5--6):} Final evaluation. This winter-season data provides a natural domain shift from the summer training data, testing VABQ robustness.
\end{itemize}

\subsection{Baseline Configurations}
\label{sec:exp_baselines}

Five configurations are compared, as summarised in Table~\ref{tab:baselines}.

\begin{table}[H]
\centering
\caption{Experimental configurations for comparative evaluation.}
\label{tab:baselines}
\begin{tabular}{@{}clll@{}}
\toprule
\textbf{ID} & \textbf{Sensor Quantisation} & \textbf{Inference Quantisation} & \textbf{Description} \\
\midrule
B1 & 16-bit uniform & FP32 & Full-precision baseline \\
B2 & 16-bit uniform & INT8 PTQ & Inference-only PTQ~\cite{smoothquant2024} \\
B3 & 8-bit uniform & FP32 & Sensor-only static reduction \\
B4 & Adaptive (fixed $\theta$) & FP32 & Adaptive threshold~\cite{atitallah2023autonomous} \\
\textbf{P} & \textbf{4/8/16-bit VABQ} & \textbf{INT8 PTQ} & \textbf{Proposed dual-layer} \\
\bottomrule
\end{tabular}
\end{table}

B1 represents the energy upper bound and accuracy upper bound. B2 isolates the effect of inference-only quantisation. B3 isolates sensor-level uniform bit reduction. B4 implements the adaptive threshold method of Atitallah et al.~(2023)~\cite{atitallah2023autonomous} with their recommended Kalman-based threshold adaptation.

\subsection{Evaluation Metrics}
\label{sec:exp_metrics}

\textbf{Signal Fidelity.} Normalised Root Mean Square Error (NRMSE) between reconstructed and original signals per channel:
\begin{equation}
    \text{NRMSE}_c = \frac{\sqrt{\frac{1}{T}\sum_{t=1}^{T}(x_c(t) - \hat{x}_c(t))^2}}{x_{c,\max} - x_{c,\min}}
    \label{eq:nrmse}
\end{equation}

\textbf{Classification Performance.} Overall accuracy, macro-averaged F1-score, and weighted F1-score for the eight-class activity recognition task.

\textbf{Energy Metrics.} Total sensor energy $E_{\text{sensor}}$, inference energy $E_{\text{infer}}$, transmission energy $E_{\text{tx}}$, and total system energy $E_{\text{total}}$, all normalised to B1.

\textbf{Bit-Width Distribution.} Percentage of measurement time at each bit-width (4, 8, 16) per channel, reflecting VABQ's adaptive behaviour.

\subsection{Implementation Details}
\label{sec:exp_implementation}

The 1D-CNN is implemented in TensorFlow 2.12 and quantised using TensorFlow Lite's PTQ pipeline with min-max calibration~\cite{nagel2021whitepaper}. The VABQ sensor-level controller is implemented in C on the ESP32 using the ESP-IDF v5.1 framework. Rolling statistics use Welford's online algorithm for numerical stability. Entropy histograms use 16 bins per channel.

Energy measurements are validated using an INA219 current sensor in series with the ESP32 power supply, sampling at 100~Hz. The measured active-mode current of 742~mA at 5~V (3.71~W) is within 4.5\% of the datasheet specification of 760~mA~\cite{esp32power2025, karmakar2024exploring}.

All experiments are repeated five times with different random seeds for model training. Reported values are means $\pm$ standard deviations. Statistical significance of accuracy differences is assessed using paired $t$-tests at $p < 0.05$.
