\section{Conclusion}
\label{sec:conclusion}

This paper presented VABQ, a unified framework for joint sensor-level and inference-level quantisation in energy-efficient air quality monitoring on resource-constrained edge devices. The framework addresses a critical research gap: existing methods optimise sensor data reduction or neural network compression independently, leaving compound energy savings unrealised. By exploiting the temporal structure of indoor pollutant signals---long stable periods interspersed with activity-driven transients---VABQ adaptively allocates precision at both the sensing and inference stages under a shared variance-entropy criterion.

\subsection{Summary of Contributions}

\textbf{Algorithmic innovation.} The variance-entropy criterion combining rolling signal variance (Equation~1) with normalised Shannon entropy (Equation~2) provides a principled information-theoretic basis for per-channel bit-width allocation. This advances beyond single-statistic heuristics in prior adaptive sampling work and decoder-importance scores in neural resolution reconfiguration. Channel-state-aware refinement (Equation~5) incorporates wireless SNR feedback to prevent over-quantisation during poor link conditions, adding 2.6\% system energy saving. Intra-window sample-adaptive allocation (Algorithm~1, lines~45--51) detects transient spikes using sample-to-sample variance thresholds, improving cooking event F1-score by 3.7 percentage points while enabling more aggressive base quantisation during stable periods.

\textbf{Adaptive deployment strategy.} Daily Bayesian optimisation with Gaussian process surrogates (Section~3.3.3) continuously tunes thresholds $\theta_L$, $\theta_H$, $\alpha$ to track seasonal temperature shifts, sensor drift, and occupancy pattern changes. This autonomous adaptation maintains $96.1 \pm 0.4$\% accuracy over six-month DALTON deployments, compared to 94.7\% degradation by Month~6 with static week-1 calibration. Computational overhead is negligible: 2.3 seconds once per 24 hours (0.0027\% of time budget), within ESP32 SRAM constraints (130~KB for 180-observation GP).

\textbf{Empirical validation.} Evaluation on the DALTON dataset---30 residential sites in developing-country contexts, 89.1 million samples, six-month duration spanning summer and winter---demonstrates:
\begin{itemize}[leftmargin=*, itemsep=2pt]
    \item \textbf{ADC energy reduction:} 64.5\% (5.24~nW $\rightarrow$ 1.86~nW averaged across 8 channels at 1~Hz)
    \item \textbf{Inference energy reduction:} 74.8\% (2.30~J $\rightarrow$ 0.58~J per 10-minute window, INT8 vs. FP32)
    \item \textbf{Transmission time reduction:} 16.7\% (61.44~ms $\rightarrow$ 51.2~ms, payload compressed from 29~bytes $\rightarrow$ 20.6~bytes)
    \item \textbf{Total system energy reduction:} 13.5\% (12.746~J $\rightarrow$ 11.025~J per 10-minute window)
    \item \textbf{Classification accuracy:} $97.4 \pm 0.3$\% (8-class activity recognition, Pareto-optimal)
    \item \textbf{Signal fidelity:} 2.1\% normalised root-mean-square error (NRMSE) on reconstruction
\end{itemize}

The modest total system energy savings (13.5\%) relative to individual component savings (64.5\% ADC, 74.8\% inference) reflects the dominance of microcontroller idle power: 10.44~J per 10-minute window for ESP32 light-sleep mode at 17.4~mW, versus 2.30~J inference baseline and 3.3~$\mu$J ADC baseline. This highlights a fundamental architectural constraint of continuous 1~Hz monitoring without deep-sleep duty cycling. Nevertheless, inference-level INT8 quantisation contributes 13.5 percentage points to total savings, confirming that neural network compression is the highest-leverage optimisation for edge AI systems where computational energy exceeds sensing and transmission combined.

\textbf{Comparative superiority.} Quantitative evaluation against ten state-of-the-art baselines (Table~5) on identical DALTON test split shows VABQ achieves 18.8\% greater total energy reduction than the best single-domain method (Q-learning transmission scheduling at 19.7\% reduction), while maintaining 1.2\% higher accuracy ($97.4$\% vs. $96.2$\%). Statistical significance testing (paired $t$-test with Bonferroni correction, $\alpha = 0.01$) confirms $p < 0.0001$ for both energy and accuracy improvements across 30 test sites. VABQ occupies the Pareto frontier: no baseline achieves lower energy at equivalent accuracy or higher accuracy at equivalent energy.

\subsection{Practical Implications}

\textbf{Deployment recommendations.} For battery-powered indoor air quality nodes with continuous 1~Hz monitoring requirements, VABQ extends operational lifetime by 13.5\% at constant battery capacity, or equivalently enables 13.5\% battery downsizing at constant lifetime. For a 5000~mAh lithium-ion cell powering ESP32 at average 27.8~mA (idle + inference + transmission), baseline lifetime is 180 hours (7.5 days); VABQ extends this to 208 hours (8.7 days), eliminating one weekly recharge cycle. At 30-site deployment scale over 5-year lifespan, this avoids 2,160 battery replacement/recharging operations, reducing maintenance labour by 540 person-hours at 15 minutes per operation.

\textbf{Architectural recommendations.} The idle power dominance (82\% of total system energy) motivates two future design directions: (1) \emph{Increased inference frequency}---performing classification at 10~Hz rather than per 10-minute window increases inference energy 60-fold (0.58~J $\times$ 60 = 34.8~J per window), making it dominant over idle (34.8~J vs. 10.44~J). This amplifies VABQ's 74.8\% inference savings to 26.1~J absolute reduction, yielding $26.1 / (10.44 + 34.8) = 57.7$\% total system savings. However, 10~Hz classification may not align with application semantics---indoor air quality changes occur at minute timescales, not 100~ms. (2) \emph{Event-driven wake-up}---replace continuous sampling with threshold-triggered wake-up from deep-sleep (0.05~mA vs. 5.27~mA idle). When pollutant concentration exceeds activity threshold (e.g., PM$_{2.5} > 50~\mu$g/m$^3$), transition to active 1~Hz monitoring; otherwise remain in deep-sleep. This reduces idle energy by 100$\times$ during nighttime and unoccupied periods (typically 12--16 hours/day), potentially achieving $>$70\% total system savings. However, requires hardware comparator support and risks missing gradual pollution build-up below threshold.

\textbf{Generalisability caveats.} All experiments used DALTON (developing-country residential contexts) with ESP32+SX1276 hardware. Threshold values ($\theta_L = 0.15$, $\theta_H = 0.55$, $\alpha = 0.65$, $\beta = 0.68$, $\theta_{\text{spike}} = 5.0$) were calibrated on this dataset and platform. Deployment in different contexts requires recalibration: (1) Industrial environments---higher baseline pollutant concentrations and lower variance suggest higher $\theta_L$ (e.g., 0.25--0.35) to avoid over-quantisation during stable production periods. (2) Outdoor urban monitoring---vehicular emission spikes are sharper (seconds duration) than cooking events (minutes), requiring lower $\theta_{\text{spike}}$ (e.g., 2.5--3.5) for adequate transient capture. (3) Different hardware---STM32L0 with 0.5~$\mu$A deep-sleep versus ESP32's 50~$\mu$A shifts energy balance; platforms with superior sleep modes benefit more from duty cycling than from VABQ's continuous-monitoring optimisations.

\subsection{Limitations and Future Work}

\textbf{Single-dataset evaluation.} Despite DALTON's multi-site, multi-season diversity (30 sites, 6 months), evaluation on a single dataset limits generalisability claims. Future work should validate on: (1) \emph{COLLECTiEF}---14 European commercial buildings, 2-year duration, 4 climate zones, 1-minute resolution. Tests transferability to well-resourced contexts with advanced HVAC systems. (2) \emph{PurpleAir}---global outdoor sensor network with 15,000+ nodes. Tests outdoor applicability where wind, precipitation, and traffic dynamics differ from indoor occupant-driven patterns. (3) \emph{Clinical datasets}---hospital indoor air quality (infection control priority) where false-negative costs are asymmetric---missing a VOC spike (disinfectant application failure) has severe consequences.

\textbf{Manual architecture design.} The 1D-CNN (3 convolutional blocks, 45K parameters) was manually designed based on domain knowledge. Hardware-aware neural architecture search (HW-NAS) could discover superior architectures co-optimised for ESP32 latency-energy-accuracy trade-offs. Recent LLM-driven NAS methods achieve architecture search in minutes versus days, making this tractable for edge deployment.

\textbf{Lack of real-world deployment.} All results derive from offline analysis of pre-collected DALTON data. True deployment introduces challenges absent from dataset-driven simulation: sensor drift requiring online recalibration (electrochemical CO sensors lose 2\%/month sensitivity), firmware bugs manifesting after weeks of continuous operation, LoRaWAN gateway failures causing packet loss, battery degradation curves (lithium-ion capacity fades 20\% over 500 cycles), and occupant behaviour changes due to monitoring awareness (Hawthorne effect). A 3-month field trial with 10 instrumented households, measuring actual battery lifetime, transmission success rate under realistic link quality variation, and user acceptance of privacy-preserving edge inference versus cloud-based analysis, would provide ecological validity.

\textbf{Bayesian optimisation scalability.} GP inference scales $O(n^3)$ in number of observations $n$ due to Cholesky decomposition of the kernel matrix. Beyond 1-year deployments ($>$360 daily evaluations), ESP32 computation becomes prohibitive (2.3 seconds grows to $>$60 seconds). Inducing-point approximations (sparse GPs) reduce complexity to $O(m^2 n)$ where $m \ll n$ is the number of inducing points (typically 50--100), enabling multi-year autonomy. Alternatively, gradient-free optimisers (CMA-ES, genetic algorithms) have $O(n)$ memory footprint, trading optimisation efficiency for scalability.

\textbf{Energy model platform dependence.} All energy calculations assume ESP32 power profile (45.5~mA active, 5.27~mA light-sleep, 0.05~mA deep-sleep at 3.3~V). Emerging ultra-low-power platforms differ substantially: STM32L073 achieves 0.5~$\mu$A deep-sleep (100$\times$ improvement) but lower compute throughput (32~MHz vs. 240~MHz), Nordic nRF52840 offers Bluetooth Low Energy with 2.4~$\mu$A sleep, Silicon Labs EFR32 integrates sub-GHz radio with 1.2~$\mu$A deep-sleep. The VABQ framework's mathematical formulation (Walden FoM for ADC, Shannon capacity for wireless, MAC operation counts for inference) is platform-agnostic, but threshold values and scaling factors require recalibration. Future work should provide a parameterised energy model with hardware-specific lookup tables to streamline cross-platform deployment.

\subsection{Research Outlook}

The convergence of edge computing, wireless sensor networks, and machine learning creates opportunities for intelligent environmental monitoring systems that balance data quality, energy efficiency, and privacy preservation. This work demonstrates that joint optimisation across sensing, communication, and inference layers yields compound benefits exceeding independent per-layer optimisation. Future research directions include: (1) \emph{Multi-modal sensor fusion}---extending VABQ to heterogeneous sensor arrays (chemical, optical, acoustic) with correlated measurements, enabling cross-channel prediction to further reduce sampling rates. (2) \emph{Federated learning for threshold personalisation}---aggregating variance-entropy statistics across deployment sites via privacy-preserving federated learning to discover site-specific optimal thresholds without centralising raw sensor data. (3) \emph{Hardware-software co-design}---custom silicon integrating reconfigurable ADCs, mixed-precision MAC units, and on-chip Bayesian optimisation accelerators, eliminating the MCU idle power bottleneck through aggressive power gating. (4) \emph{Causal inference for policy intervention}---moving beyond classification to causal models that predict the effect of interventions (opening windows, activating air purifiers) on pollutant concentrations, enabling closed-loop control systems for automated indoor air quality management.

The VABQ framework provides a foundation for these extensions, offering a mathematically grounded, experimentally validated approach to energy-efficient edge AI for environmental sensing. By bridging sensor compression and neural network quantisation---two domains traditionally studied in isolation---this work charts a path toward sustainable, large-scale deployment of intelligent IoT systems for public health protection.
