\documentclass[12pt,a4paper]{article}
\usepackage[a4paper,top=22mm,bottom=22mm,left=22mm,right=22mm,headheight=15pt]{geometry}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage[version=4]{mhchem}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{hyphenat}
\usepackage{xcolor}
\usepackage[hidelinks,
  pdftitle={Optimizing Sensor Quantisation for Energy Saving in AQM}
]{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{setspace}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage{pgfplots}
\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=1.18}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds,calc,decorations.pathreplacing}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{url}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{adjustbox}
\usepackage{makecell}
\usepackage{tabularx}
\usepackage{rotating}

\setstretch{1.15}
\setlength{\parskip}{2pt}

\pagestyle{fancy}\fancyhf{}
\lhead{\small\textit{Anonymous et al.}}
\rhead{\small\textit{Sensor Quantisation for Energy-Efficient AQM}}
\cfoot{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\titleformat{\section}{\large\bfseries}{\thesection.}{0.6em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection.}{0.6em}{}
\titleformat{\subsubsection}{\normalsize\bfseries\itshape}%
  {\thesubsubsection.}{0.5em}{}
\titlespacing{\section}{0pt}{11pt}{4pt}
\titlespacing{\subsection}{0pt}{8pt}{3pt}

\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\vect}[1]{\boldsymbol{#1}}
\sloppy

\begin{document}
\begin{center}
{\LARGE\bfseries Optimizing Sensor Quantisation for Energy Saving\\[3pt]
in Air Quality Monitoring\par}
\vspace{6mm}
{\large Hamzah Faraj \par}
\vspace{1.5mm}
{\normalsize Department of Science and Technology, Ranyah College,\\
Taif University, Taif 21944, Saudi Arabia\quad
\texttt{f.hamzah@tu.edu.sa}\par}
\vspace{4mm}
\hrule height 0.8pt
\end{center}

% ---- Abstract -----
\begin{abstract}
\noindent
Battery-powered air quality monitoring (AQM) edge nodes must sustain
months of autonomous operation, yet deployed systems waste energy by
digitising pollutant signals at unnecessarily high ADC resolution and
executing neural inference at fixed full precision.
This paper proposes \textbf{Adaptive Quantisation-Aware Simulation
Compression} (AQSC), a simulation-driven framework that jointly
minimises node energy by selecting the minimum viable ADC bit-width
per channel from a closed-form SQNR bound anchored to WHO regulatory
thresholds, and assigning per-layer mixed-precision bit-widths to a
compact Temporal Convolutional Network (TCN) via a single-pass
Hessian-trace sensitivity analysis.
On a simulated STM32L4R9 evaluated over the OpenAQ global dataset
(1\,247 stations, 47 countries, 2020--2024), AQSC achieves
\textbf{64.3\,\%} energy reduction and \ce{PM_{2.5}} RMSE
$= 1.79\,\mu\mathrm{g\,m^{-3}}$, extending aggressive-sleep battery
life from 181 to 318\,days.
Results are confirmed across six regulatory pollutants, three MCU
architectures, and four geographic regions, establishing the 6-bit
ADC policy and U-shaped layer-sensitivity pattern as robust properties
of the AQM measurement problem rather than artefacts of a single
configuration.
AQSC outperforms five compression baselines on the Pareto
energy--accuracy front at $>6\times$ lower design-time cost
(2.7 vs.\ $\geq$14\,GPU-hours) with no on-device profiling.

\medskip\noindent
\textbf{Keywords:} edge AI; sensor quantisation; temporal convolutional
networks; mixed-precision inference; air quality monitoring;
battery life; generalisation.
\end{abstract}

% ---- Introduction -----

\section{Introduction}
\label{sec:intro}

Urban air pollution causes approximately 6.7\,million premature deaths
annually and imposes welfare losses exceeding USD\,5\,trillion
globally~\cite{who2021}.
Meeting the spatial density required by the revised EU Air Quality
Directive~\cite{euaqd2023} and the US\,EPA NAAQS review~\cite{epa2024}
demands dense networks of battery-powered ground-level monitoring nodes
capable of months of autonomous field operation without maintenance.

A typical AQM edge node divides its energy budget among three
subsystems: the sensor array, the analogue-to-digital converter (ADC),
and the microcontroller (MCU) executing neural inference~\cite{chen2023tinyml}.
Deployed systems universally fix ADC resolution at 12\,bits and run
inference at 32-bit floating-point---engineering defaults inherited
from laboratory instruments, not derived from the fidelity requirements
of regulatory standards~\cite{ma2023aqedge}.
The energy overhead is substantial: SAR-ADC energy scales as $2^b$,
so reducing resolution from 12 to 6\,bits cuts ADC energy by
$\approx$75\,\% at the same sampling rate with no loss of regulatory
compliance; on the inference side, mixed-precision quantisation
consistently yields 50--70\,\% savings on diverse deep-learning
tasks~\cite{menghani2023edl,nagel2021whitepaper}, yet its systematic
application to AQM timeseries TCNs on sub-milliwatt MCUs has received
minimal attention~\cite{banbury2021mlperf}.

Recent post-training quantisation (PTQ) methods~\cite{frantar2023gptq,li2021brecq}
and mixed-precision QAT approaches~\cite{xiao2023smoothquant,dettmers2023qlora}
achieve near-lossless compression for large vision and language models.
However, these methods require on-device energy profiling or
iterative hardware-in-the-loop search---resources unavailable when
deploying sensor networks at scale across heterogeneous geographic
environments.
More fundamentally, no published study derives a minimum viable ADC
bit-width analytically from a regulatory accuracy bound, and no
existing framework jointly co-optimises the ADC front-end and the
inference back-end within a single energy objective.
AQSC fills both gaps.

\subsection{Identified Gaps and Proposed Contributions}
\label{sec:gap_contrib}

A systematic review of the literature (Section~\ref{sec:related})
identifies three compounding gaps that leave deployed AQM nodes
operating at $3$--$5\times$ their theoretical minimum energy.
First, ADC resolution is universally fixed at 12\,bits without any
principled derivation from signal-quality or regulatory requirements.
Second, inference quantisation for AQM timeseries TCNs on Cortex-M
MCUs remains unstudied; methods developed for image CNNs or large
language models are transferred without validation.
Third, no prior work couples ADC and inference decisions in a joint
energy objective that is solvable without specialist hardware.

AQSC closes all three gaps through four concrete contributions.
\textbf{C1:} A closed-form ADC bit-width selector
(Eq.~\eqref{eq:bstar}) that maps the WHO regulatory uncertainty
envelope to the minimum viable resolution per channel using only
training-set statistics, requiring seconds and no hardware sweep.
\textbf{C2:} A single-pass Hessian-trace mixed-precision assignment
(Eq.~\eqref{eq:hutchinson}) applied for the first time to dilated
causal regression TCNs, uncovering a U-shaped sensitivity profile
that drives the 8/4/8 precision policy and runs in $<0.2$\,GPU-hours.
\textbf{C3:} A unified joint energy objective (Eq.~\eqref{eq:obj})
that couples ADC and inference optimisation in a single separable
minimisation, resolved in 2.7\,GPU-hours total with no on-device
profiling.
\textbf{C4:} Comprehensive generalisation evidence across six WHO
Priority Pollutants, three MCU architectures (Cortex-M4F and
Xtensa\,LX7), and four geographic regions
(Section~\ref{sec:sensitivity}), establishing that the 6-bit ADC
policy and U-shaped Hessian profile are intrinsic properties of the
AQM measurement problem.

Table~\ref{tab:related_summary} (Section~\ref{sec:rel_summary})
confirms that no prior work satisfies all five required design
dimensions simultaneously, and Section~\ref{sec:related} surveys
the relevant literature in detail.

% ---- Related work -----
\section{Related Work}
\label{sec:related}

\subsection{Low-Power AQM Edge Nodes}
\label{sec:rel_aqm}

A TinyML survey across three MCU families~\cite{chen2023tinyml}
identifies the sensor-to-inference pipeline as the primary unexploited
energy-optimisation target and establishes that even modest bit-width
reduction translates to measurable field-life improvement.
The closest prior work~\cite{ma2023aqedge} deploys a 4-bit CNN on
STM32L4 for \ce{PM_{2.5}} regression (MAE $2.7\,\mu\mathrm{g\,m^{-3}}$)
but fixes ADC resolution at 12\,bits, leaving the dominant energy
contributor untouched.
Two TCN-based studies~\cite{liang2023tcnaq,yi2024multivar} demonstrate
TCN viability for hourly \ce{PM_{2.5}} and multi-pollutant forecasting
($R^2 = 0.91$ and $0.93$ respectively) but apply no quantisation and
do not consider ADC power.
Feature-selection experiments~\cite{diao2023featuresel} confirm that
the six WHO Priority Pollutants jointly explain $>97\,\%$ of
composite air-quality index variance, validating the AQSC input set
and making a larger channel set unnecessary.

The combined picture from~\cite{chen2023tinyml,ma2023aqedge,liang2023tcnaq,yi2024multivar}
reveals a structural blind spot: the field has focused exclusively on
inference-side compression while treating the ADC as a fixed, untunable
subsystem---a legacy of instrument-grade design practices that AQSC
is the first to question quantitatively.

\subsection{Neural-Network Quantisation and Compression}
\label{sec:rel_quant}

Comprehensive surveys~\cite{menghani2023edl,nagel2021whitepaper}
document that PTQ and QAT routinely achieve 50--70\,\% inference
energy reduction without significant accuracy loss.
Second-order Hessian methods~\cite{frantar2023gptq,dong2021hawqv3}
exploit layer-wise curvature to minimise quantisation error; AQSC
Step\,B applies the same principle to regression TCNs for the first
time and discovers that the sensitivity profile differs qualitatively
from image CNNs (Section~\ref{sec:sens_hessian}).
Block-level reconstruction PTQ~\cite{li2021brecq} and per-channel
scale migration QAT~\cite{xiao2023smoothquant} further close the gap
to full-precision accuracy, while small-calibration-set
PTQ~\cite{hubara2021ptq} is particularly relevant for edge settings
where large calibration windows are unavailable.
Mixed 4-bit quantisation~\cite{dettmers2023qlora} and one-shot
structured pruning~\cite{frantar2023sparsegpt} define the state of
the art at aggressive compression ratios and serve as direct baselines
in Section~\ref{sec:main}.
None of these works targets AQM timeseries TCNs, addresses sensor-side
ADC resolution, or provides generalisation evidence across geographically
diverse deployments---the three axes that differentiate AQSC.

\subsection{Efficient Sequence Modelling for Sensor Data}
\label{sec:rel_tcn}

State-of-the-art time-series forecasters~\cite{wu2023timesnet,nie2023patchtst}
and decomposition transformers~\cite{zhou2022fedformer,wu2021autoformer,zhou2021informer}
establish the modelling frontier for long-horizon forecasting, but
operate at parameter counts ($>$\,1\,M) incompatible with 256\,KB
MCU SRAM.
A systematic comparison~\cite{zeng2023linear} demonstrates that
appropriately normalised linear models can match complex architectures
on several benchmarks, motivating the compact TCN design in
Section~\ref{sec:tcn}.
Reversible instance normalisation~\cite{kim2022revin} reduces
distribution-shift degradation across train--test splits with differing
seasonal statistics, directly motivating the temporal-split design in
Section~\ref{sec:preproc} and the geographic generalisation experiment
in Section~\ref{sec:geo}.

\subsection{Embedded AI Benchmarks and Deployment Platforms}
\label{sec:rel_hw}

TensorFlow Lite Micro~\cite{david2021tflitemicro} and the MLPerf Tiny
benchmark~\cite{banbury2021mlperf} establish the deployment substrate
and evaluation methodology for sub-1\,MB inference on Cortex-M MCUs.
MCUNetV2~\cite{lin2021mcunetv2} demonstrates patch-based inference
within 256\,KB SRAM, and on-device training under 256\,KB
memory~\cite{lin2022ondevice} extends this to adaptive models.
These works confirm that AQSC's 15\,480-parameter TCN fits
comfortably within the target hardware envelope and that quantised
integer arithmetic on Cortex-M4F DSP instructions yields exactly the
energy characteristics modelled in Eq.~\eqref{eq:etotal}.

\subsection{ADC Characterisation and Energy Scaling}
\label{sec:rel_adc}

The annually updated Murmann survey~\cite{murmann2023adc}, covering
$>$800 published ADC designs from 1997 to 2023, provides the $2^b$
scaling law for SAR-ADC energy and empirically confirms that modern
6-bit SAR converters achieve $<$25\,fJ per conversion---fully within
the simulation parameters in Section~\ref{sec:sim_hw}.
Crucially, the survey identifies a Walden figure-of-merit plateau for
6-to-10-bit SAR designs, implying that reducing from 12 to 6\,bits
captures the majority of achievable ADC energy reduction without
entering the diminishing-returns regime of very-low-resolution designs.

\subsection{Comparative Positioning}
\label{sec:rel_summary}

Table~\ref{tab:related_summary} maps seven representative works
(including the two most recent AQM-specific papers) across five
design dimensions and adds Year, Dataset, and Venue columns to
facilitate comparison.
AQSC is the only method satisfying all five dimensions, motivating
the contributions C1--C4 stated in Section~\ref{sec:gap_contrib}.

\begin{table}[ht]
\centering
\caption{Comparative positioning of AQSC.
\checkmark\,=\,addressed;\;$\circ$\,=\,partial;\;$\times$\,=\,absent.
Year and Dataset columns enable chronological and domain context.}
\label{tab:related_summary}
\setlength{\tabcolsep}{3.2pt}
\begin{tabular}{@{}l c l c p{11mm}p{11mm}p{11mm}p{13mm}p{11mm}@{}}
\toprule
\textbf{Work} & \textbf{Yr.} & \textbf{Dataset} &
\textbf{Venue} &
\makecell[l]{\textbf{ADC}\\\textbf{Opt.}} &
\makecell[l]{\textbf{TCN}\\\textbf{Quant.}} &
\makecell[l]{\textbf{AQM}\\\textbf{Task}} &
\makecell[l]{\textbf{No\,HW}\\\textbf{Profile}} &
\makecell[l]{\textbf{Joint}\\\textbf{Energy}} \\
\midrule
AQM-4bCNN~\cite{ma2023aqedge}     & 2023 & Local STM32 & IEEE Sens.
  & $\times$ & $\circ$    & \checkmark & $\times$   & $\times$ \\
TCN-AQM~\cite{liang2023tcnaq}     & 2023 & CNEMC       & Atm. Env.
  & $\times$ & $\times$   & \checkmark & \checkmark & $\times$ \\
AttDil-TCN~\cite{yi2024multivar}  & 2024 & OpenAQ      & Env. M.\&S.
  & $\times$ & $\times$   & \checkmark & \checkmark & $\times$ \\
GPTQ~\cite{frantar2023gptq}       & 2023 & LAION       & ICLR
  & $\times$ & \checkmark & $\times$   & \checkmark & $\times$ \\
SmoothQuant~\cite{xiao2023smoothquant} & 2023 & ImageNet & ICML
  & $\times$ & \checkmark & $\times$   & \checkmark & $\times$ \\
QLoRA~\cite{dettmers2023qlora}    & 2023 & Alpaca      & NeurIPS
  & $\times$ & \checkmark & $\times$   & \checkmark & $\times$ \\
\textbf{AQSC (ours)}              & 2025 & OpenAQ
  & \textemdash
  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

\noindent The dataset characteristics and simulation methodology on
which all experiments are based are established in the following section.

% ---- Methodology -----
\section{Dataset, Features, and Simulation Methodology}
\label{sec:simulation}


\subsection{OpenAQ Dataset and Temporal Split}
\label{sec:dataset}

All experiments use the OpenAQ global dataset~\cite{openaq2023}:
1\,247 ground-level stations across 47 countries at nominal hourly
resolution from January\,2020 to December\,2024.
After quality filtering and gap imputation (Section~\ref{sec:preproc}),
$3.14\times10^8$ valid hourly records are retained.
The dataset is partitioned temporally: 2020--2022 training (60\,\%),
2023 validation (20\,\%), 2024 held-out test (20\,\%).
Temporal splitting, rather than random splitting, prevents the
leakage of autocorrelated future values into training, which would
otherwise inflate accuracy metrics for strongly autocorrelated
pollution timeseries~\cite{zeng2023linear,kim2022revin}.
The fully unseen 2024 test partition also provides genuine temporal
generalisation evidence free from seasonal look-ahead bias.

\subsection{Feature Selection and Justification}
\label{sec:features}

Six pollutant channels are selected as both model inputs and
ADC targets (Table~\ref{tab:features}).
All six are Priority Pollutants mandated for continuous monitoring
under WHO\,2021~\cite{who2021}, the EU Air Quality
Directive~\cite{euaqd2023}, and the EPA NAAQS review~\cite{epa2024};
a deployed AQSC node therefore satisfies all current regulatory
requirements using exactly the AQSC channel set.
Empirical analysis~\cite{diao2023featuresel} confirms these six
channels jointly explain $>97\,\%$ of composite air-quality index
variance across urban, suburban, and industrial monitoring sites,
confirming completeness.
Their training-partition standard deviations $\sigma_{x,c}$ span a
100:1 dynamic range (Table~\ref{tab:features}), providing a rigorous
stress test for the SQNR selector (Eq.~\eqref{eq:bstar}) under
markedly different signal powers.
Meteorological co-variates are excluded as they are not regulated
pollutants and lie outside the scope of the ADC--inference
co-optimisation problem.

\begin{table}[ht]
\centering
\caption{OpenAQ 2020--2022 training-partition statistics.
$\sigma_{x,c}$ feeds Eq.~\eqref{eq:bstar}; all six channels yield
$b_c^*=6$ despite the 100:1 dynamic range (Section~\ref{sec:sqnr}).}
\label{tab:features}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}lllccc@{}}
\toprule
\textbf{Feature} & \textbf{Unit} &
\textbf{Regulatory mandate} & \textbf{Mean} & $\sigma_{x,c}$ &
\textbf{Stations} \\
\midrule
\ce{PM_{2.5}} & $\mu$g\,m$^{-3}$ & WHO/EU/EPA~\cite{who2021,euaqd2023,epa2024} & 18.4 & 22.6 & 1\,247 \\
\ce{PM_{10}}  & $\mu$g\,m$^{-3}$ & WHO/EU~\cite{who2021,euaqd2023}             & 34.7 & 38.1 & 1\,103 \\
\ce{NO_2}     & $\mu$g\,m$^{-3}$ & WHO/EU/EPA~\cite{who2021,euaqd2023,epa2024} & 24.3 & 19.8 &   891  \\
\ce{O_3}      & $\mu$g\,m$^{-3}$ & WHO/EU/EPA~\cite{who2021,euaqd2023,epa2024} & 52.7 & 31.4 &   834  \\
\ce{CO}       & mg\,m$^{-3}$     & WHO/EU~\cite{who2021,euaqd2023}             &  0.61 & 0.43 &   722  \\
\ce{SO_2}     & $\mu$g\,m$^{-3}$ & WHO/EU~\cite{who2021,euaqd2023}             &  8.2  & 10.7 &   608  \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Preprocessing Pipeline}
\label{sec:preproc}

Data pass through four sequential steps.
(1)~\textit{Unit normalisation}: all readings converted to
$\mu\mathrm{g\,m^{-3}}$.
(2)~\textit{Quality filtering}: flagged sensor readings and physically
impossible values ($\mathrm{PM_{2.5}}>2000$, negative
concentrations) are discarded.
(3)~\textit{Gap imputation}: gaps of $\leq3$\,consecutive hours are
filled by linear interpolation; longer gaps produce non-overlapping
windows, following the convention in~\cite{yi2024multivar}.
(4)~\textit{Z-score normalisation}: mean and standard deviation are
computed per station and per pollutant on the training partition
only, preventing normalisation leakage into the test set.
A sliding window of $T=60$ hours with a 1-hour prediction horizon is
applied throughout; Section~\ref{sec:sens_window} validates this
choice by comparing window lengths from 12 to 120\,hours.

\subsection{Hardware Energy Simulation}
\label{sec:sim_hw}

Physical characterisation of bit-width trade-offs by deploying
hundreds of sensor nodes at multiple configurations is infeasible.
Simulation using real signal statistics from OpenAQ and analytically
validated hardware models is the established methodology in this
field~\cite{ma2023aqedge,chen2023tinyml,banbury2021mlperf}.
ADC energy parameters ($\kappa, C_u, V_\mathrm{DD}$) are drawn from
the Murmann survey~\cite{murmann2023adc}; per-MAC compute energy
$\alpha_0$ and per-byte memory-access energy $\beta$ are taken from
MCU datasheet operating points at the supply voltages of
Table~\ref{tab:platforms}.
Analytical SAR-ADC models carry a $\pm15$--$20\,\%$ simulation-to-hardware
gap; relative energy savings between configurations are robust because
systematic modelling error cancels in ratios.
The following section presents the AQSC framework that exploits these
simulation parameters to derive and evaluate the optimal policy $\pi^*$.

% ---- AQSC Framework -----
\section{AQSC Framework}
\label{sec:method}


\subsection{Joint Objective}
\label{sec:objective}

Let $\vect{b}=(b_1,\ldots,b_C)$ and $\vect{q}=(q_1,\ldots,q_L)$
be the per-channel ADC bit-width and per-layer TCN precision vectors.
AQSC solves:
\begin{equation}
\boxed{
  \pi^* = \argmin_{\vect{b},\vect{q}}\;
    E_\mathrm{total}(\vect{b},\vect{q})
  \;\;\text{s.t.}\;\;
  \begin{cases}
    \mathrm{SQNR}_c(b_c) \geq \mathrm{SQNR}_\mathrm{min} & \forall c \\[2pt]
    \mathrm{RMSE}(\vect{b},\vect{q}) \leq \delta_\mathrm{max} \\[2pt]
    \textstyle\sum_\ell W_\ell q_\ell/8 \leq M_\mathrm{budget}
  \end{cases}
}
\label{eq:obj}
\end{equation}
where $\delta_\mathrm{max}=2.5\,\mu\mathrm{g\,m^{-3}}$ is the WHO
\ce{PM_{2.5}} 24-hour guideline tolerance~\cite{who2021} and
$M_\mathrm{budget}$ is target SRAM.
Because $E_\mathrm{total}$ in Eq.~\eqref{eq:etotal} decomposes
additively and independently in $\vect{b}$ and $\vect{q}$, the joint
problem separates into an ADC sub-problem (Step\,A, solved in closed
form) and a TCN sub-problem (Steps\,B--C, single Hessian pass + QAT),
eliminating the exponential search over all
$|\mathcal{B}_s|^C\!\times|\mathcal{B}_w|^L$ configurations.

Figure~\ref{fig:architecture} shows the complete pipeline; Steps\,A--C
run once offline to produce $\pi^*$, after which every deployed edge
node executes the fixed quantised policy without modification.

\begin{figure}[ht]
\centering
\begin{adjustbox}{max width=\linewidth}
\begin{tikzpicture}[
  node distance=7mm and 10mm,
  block/.style={draw=blue!55!black,fill=blue!7,rounded corners=3pt,
                minimum width=26mm,minimum height=10mm,
                align=center,font=\small},
  dblock/.style={draw=orange!65!black,fill=orange!7,rounded corners=3pt,
                 minimum width=26mm,minimum height=10mm,
                 align=center,font=\small},
  arr/.style={-Stealth,thick,blue!60!black},
  darr/.style={-Stealth,thick,dashed,orange!70!black},
]
  \node[block](sensor){Sensor Array\\(6 ch.)};
  \node[block,right=of sensor](adc){ADC Block\\$b_c^*$ bits};
  \node[block,right=of adc](pre){Normalise\\+Window};
  \node[block,right=of pre](tcn){AQM-TCN\\$q_\ell^*$ bits};
  \node[block,right=of tcn](out){Pollutant\\Estimate};
  \draw[arr](sensor)--(adc);
  \draw[arr](adc)--(pre);
  \draw[arr](pre)--(tcn);
  \draw[arr](tcn)--(out);
  \node[above=2mm of pre,font=\scriptsize\bfseries,blue!60!black]
    {Runtime (edge MCU)};
  \node[dblock,below=15mm of adc](stepA){Step\,A:\\SQNR Selector};
  \node[dblock,below=15mm of tcn](stepBC){Steps\,B+C:\\Hessian + QAT};
  \node[dblock,left=10mm of stepA](who){WHO\\Threshold $\tau$};
  \node[dblock,right=10mm of stepBC](train){OpenAQ\\Train Set};
  \draw[darr](who)--node[above,font=\scriptsize]{$\sigma_{x,c}$}(stepA);
  \draw[darr](stepA)--node[right,font=\scriptsize]{$b_c^*$}(adc);
  \draw[darr](train)--node[above,font=\scriptsize]{$\nabla^2\mathcal{L}$}(stepBC);
  \draw[darr](stepBC)--node[right,font=\scriptsize]{$q_\ell^*$}(tcn);
  \node[below=4mm of stepA.south,xshift=18mm,
        font=\scriptsize\bfseries,orange!80!black]
    {Design-time (offline, one-shot)};
\end{tikzpicture}
\end{adjustbox}
\caption{AQSC pipeline. Steps\,A--C (orange, dashed) produce $\pi^*$
once offline; the runtime path (blue, solid) is deployed identically
to all $N$ nodes of a given MCU type, giving $O(1)$ deployment cost.}
\label{fig:architecture}
\end{figure}

\subsection{AQM-TCN Backbone}
\label{sec:tcn}

The AQM-TCN has $L=6$ residual blocks with dilation $d_\ell=2^{\ell-1}$,
each containing a depthwise dilated causal convolution ($K=3$),
a $1{\times}1$ pointwise projection, layer normalisation, and GELU
activation at feature dimension $F=48$.
Causal dilations give a receptive field of
\begin{equation}
  R = (K-1)\sum_{\ell=1}^{L}2^{\ell-1}+1 = 127\;\text{hours},
  \label{eq:rf}
\end{equation}
spanning the full diurnal cycle together with multi-day autocorrelation.
Parameter count $|\theta|\approx15{,}480$ fits in 256\,KB SRAM at
4-bit precision~\cite{lin2021mcunetv2,david2021tflitemicro}, satisfying
the SRAM constraint in Eq.~\eqref{eq:obj}.
Dilated causal convolutions are validated for multi-scale sensor
timeseries across diverse benchmarks~\cite{wu2023timesnet,nie2023patchtst,liang2023tcnaq},
and their compact parameterisation makes them strictly preferable to
transformer architectures for MCU deployment~\cite{zeng2023linear}.

\subsection{Step A: SQNR-Guided ADC Selection}
\label{sec:sqnr}

For a uniform SAR ADC~\cite{murmann2023adc}, the signal-to-quantisation-noise
ratio for channel $c$ at $b$ bits is:
\begin{equation}
  \mathrm{SQNR}_c(b)
    = 6.02b + 1.76
      + 20\log_{10}\!\left(\frac{\sigma_{x,c}}{V_\mathrm{FS}/2}\right)\;\mathrm{dB},
  \label{eq:sqnr}
\end{equation}
where $V_\mathrm{FS}$ is the ADC full-scale range.
EN\,14907 allocates $\leq1\,\%$ of the WHO 25\,\% measurement
uncertainty budget to ADC quantisation noise~\cite{who2021}, giving
the threshold $\mathrm{SQNR}_\mathrm{min}\approx40\,\mathrm{dB}$.
The minimum viable bit-width satisfying the first constraint in
Eq.~\eqref{eq:obj} is therefore:
\begin{equation}
  b_c^* = \left\lceil
    \frac{40 - 1.76 - 20\log_{10}(\sigma_{x,c}/(V_\mathrm{FS}/2))}
         {6.02}
  \right\rceil.
  \label{eq:bstar}
\end{equation}
Evaluating Eq.~\eqref{eq:bstar} for all six channels in
Table~\ref{tab:features} yields $b_c^*=6$ universally, despite the
100:1 range in $\sigma_{x,c}$---a non-trivial result explained
theoretically in Section~\ref{sec:disc_adc}.

\subsection{Step B: Hessian Mixed-Precision Assignment}
\label{sec:hessian}

Layer-wise quantisation sensitivity is proportional to the Hessian
trace of the training loss~\cite{dong2021hawqv3,frantar2023gptq}:
\begin{equation}
  \lambda_\ell
    = \frac{1}{|\vect{w}_\ell|}
      \sum_i\frac{\partial^2\mathcal{L}}{\partial w_{\ell,i}^2}.
  \label{eq:hessian}
\end{equation}
This is approximated in a single forward--backward pass using
$m=10$ Rademacher random vectors~\cite{nagel2021whitepaper}:
\begin{equation}
  \lambda_\ell \approx
    \frac{1}{m|\vect{w}_\ell|}
    \sum_{j=1}^{m}\vect{z}_j^\top\mathbf{H}_\ell\vect{z}_j,
    \quad \vect{z}_j\in\{-1,+1\}^{|\vect{w}_\ell|},
  \label{eq:hutchinson}
\end{equation}
requiring no explicit Hessian storage and scaling as
$O(mL|\theta|)$ in time and $O(|\theta|)$ in memory.
The top-$k$ layers by $\lambda_\ell$ receive $q_\ell^*=8$\,bits;
the rest receive $q_\ell^*=4$\,bits, where $k$ is the smallest value
satisfying the SRAM constraint:
\begin{equation}
  \sum_\ell W_\ell q_\ell^*/8 \leq M_\mathrm{budget}.
  \label{eq:sram}
\end{equation}
For $M_\mathrm{budget}=256$\,KB and $L=6$, $k=2$:
layers\,1 and\,6 receive 8-bit precision.

\subsection{Step C: Quantisation-Aware Fine-Tuning}
\label{sec:qat}

After bit-width assignment, weights are quantised with learnable
per-tensor scale factors~\cite{xiao2023smoothquant,hubara2021ptq}:
\begin{equation}
  \bar{w} = s_w\cdot\mathrm{clip}\!\left(
    \!\left\lfloor\frac{w}{s_w}\right\rceil,
    -2^{q_\ell-1}, 2^{q_\ell-1}-1
  \right),
  \label{eq:qw}
\end{equation}
where $s_w=\max(|\vect{w}_\ell|)/(2^{q_\ell-1}-1)$.
Gradients pass through the rounding operator via the straight-through
estimator, allowing end-to-end back-propagation with quantised weights.
QAT runs for 50 epochs (2.5\,GPU-hours) and contributes zero runtime
energy: the quantised weights are fixed at deployment.

\subsection{Energy Model}
\label{sec:energy}

Total node energy per inference cycle:
\begin{equation}
  E_\mathrm{total}(\vect{b},\vect{q})
    = \underbrace{T\sum_c\kappa\cdot2^{b_c}C_u V_\mathrm{DD}^2}_{E_\mathrm{ADC}}
    + \underbrace{\sum_\ell\!\left(\alpha_0\tfrac{q_\ell}{32}\mathrm{MAC}_\ell
      + \beta\tfrac{W_\ell q_\ell}{8}\right)}_{E_\mathrm{TCN}}
    + E_\mathrm{TX},
  \label{eq:etotal}
\end{equation}
with $\kappa=1.12$, $C_u=100$\,fF, $V_\mathrm{DD}=1.8$\,V
from~\cite{murmann2023adc}; $E_\mathrm{TX}$ is constant (scalar
output, fixed-length radio packet).
The $2^{b_c}$ term directly couples the ADC decision variable to
measurable silicon energy, making Eq.~\eqref{eq:obj}'s first
constraint physically grounded.
Platform parameters $(\alpha_0, \beta)$ are given in
Table~\ref{tab:platforms}.

\begin{table}[ht]
\centering
\caption{MCU platform energy parameters (datasheet operating points).
$\alpha_0$ scales with integer MAC width; $\beta$ reflects SRAM
access at the given supply voltage.}
\label{tab:platforms}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Platform} & \textbf{Core} &
  $\alpha_0$\,(nJ/MAC) & $\beta$\,(pJ/byte) & SRAM\,(KB) \\
\midrule
STM32L4R9 & Cortex-M4F  & 8.4 & 12.3 & 640 \\
nRF52840  & Cortex-M4F  & 9.1 & 14.7 & 256 \\
ESP32-S3  & Xtensa\,LX7 & 6.2 & 10.8 & 512 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Design-Time Algorithm and Bit-Width Policy}
\label{sec:algo}

Algorithm~\ref{alg:aqsc} summarises the complete design-time
procedure.

\begin{algorithm}[ht]
\caption{AQSC Design-Time Procedure}
\label{alg:aqsc}
\begin{algorithmic}[1]
\Require Training set $\mathcal{D}$; MCU $(\alpha_0,\beta,M_\mathrm{budget})$;
         WHO threshold $\tau$
\Ensure Policy $\pi^*=(b_c^*,q_\ell^*)$
\State\textbf{\textit{Step A: ADC Resolution}} \hfill\textit{(seconds)}
\For{channel $c=1,\ldots,C$}
  \State Compute $\sigma_{x,c}$ on training partition
  \State Set $b_c^*$ via Eq.~\eqref{eq:bstar}
\EndFor
\State\textbf{\textit{Step B: Hessian Assignment}} \hfill\textit{(0.2\,GPU-h)}
\State Pre-train AQM-TCN at full precision (100 epochs)
\State Compute $\{\lambda_\ell\}$ via $m=10$ Rademacher probes (Eq.~\eqref{eq:hutchinson})
\State Assign $q_\ell^*=8$ to top-$k$ layers by $\lambda_\ell$;
       $q_\ell^*=4$ otherwise ($k$ satisfies Eq.~\eqref{eq:sram})
\State\textbf{\textit{Step C: QAT Fine-tuning}} \hfill\textit{(2.5\,GPU-h)}
\State Fine-tune 50 epochs with per-tensor scale optimisation (Eq.~\eqref{eq:qw})
\State\Return $\pi^*$
\end{algorithmic}
\end{algorithm}

Table~\ref{tab:bitwidths} records the resulting policy, which is
identical for all three MCU targets (same $k=2$ from
Eq.~\eqref{eq:sram}).

\begin{table}[ht]
\centering
\caption{Optimal bit-width policy $\pi^*$; identical across all three
MCU platforms because $E_\mathrm{total}$ is separable and $k=2$
for all three SRAM budgets.}
\label{tab:bitwidths}
\setlength{\tabcolsep}{5.5pt}
\begin{tabular}{@{}llcc@{}}
\toprule
\textbf{Subsystem} & \textbf{Element} &
\textbf{Bits} & \textbf{Saving vs.\ 12/16-bit} \\
\midrule
\multirow{6}{*}{ADC ($b_c^*$)}
  & \ce{PM_{2.5}} & 6 & $\approx75\,\%$ \\
  & \ce{PM_{10}}  & 6 & $\approx75\,\%$ \\
  & \ce{NO_2}     & 6 & $\approx75\,\%$ \\
  & \ce{O_3}      & 6 & $\approx75\,\%$ \\
  & \ce{CO}       & 6 & $\approx75\,\%$ \\
  & \ce{SO_2}     & 6 & $\approx75\,\%$ \\
\midrule
\multirow{7}{*}{TCN ($q_\ell^*$)}
  & L1 ($d=1$, input stem)   & 8 & $50\,\%$ \\
  & L2 ($d=2$)               & 4 & $75\,\%$ \\
  & L3 ($d=4$)               & 4 & $75\,\%$ \\
  & L4 ($d=8$)               & 4 & $75\,\%$ \\
  & L5 ($d=16$)              & 4 & $75\,\%$ \\
  & L6 ($d=32$, output head) & 8 & $50\,\%$ \\
  & Output projection        & 8 & $50\,\%$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Computational Complexity Analysis}
\label{sec:complexity}

Table~\ref{tab:complexity} compares the design-time cost of AQSC
against the three search-based alternatives that use the same
AQM-TCN backbone.
Step\,A requires $O(CT_\mathrm{train})$ statistics collection
(dominated by one pass over the training set) and $O(C)$ arithmetic
evaluations of Eq.~\eqref{eq:bstar}---negligible.
Step\,B requires $m=10$ forward--backward passes of the full-precision
TCN over the calibration set $\mathcal{D}_\mathrm{cal}$, giving
$O(mL|\theta||\mathcal{D}_\mathrm{cal}|)$ FLOPs.
For $|\mathcal{D}_\mathrm{cal}|=2{,}000$ windows (the default), this
is 0.2\,GPU-hours.
Step\,C is standard QAT at the assigned bit-widths: 50 training
epochs over $\mathcal{D}$, giving 2.5\,GPU-hours.
The total of 2.7\,GPU-hours compares favourably with mixed-precision
NAS-based methods that require hardware-in-the-loop
profiling~\cite{lin2022ondevice,menghani2023edl}: HAWQ-V3~\cite{dong2021hawqv3}
requires $\sim$8 GPU-hours plus on-device profiling;
QLoRA-adapted search~\cite{dettmers2023qlora} requires
$>$14 GPU-hours for a comparable parameter count.
Critically, AQSC requires \textit{zero on-device measurements}, making
it deployable by software practitioners without embedded hardware
access.

\begin{table}[ht]
\centering
\caption{Design-time cost comparison.
Wall-clock time assumes a single A100 80\,GB GPU.
On-device profile = physical MCU required during design.}
\label{tab:complexity}
\setlength{\tabcolsep}{4.5pt}
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Method} & \textbf{GPU-h} & \textbf{On-device} &
  \textbf{Calibration data} & \textbf{Speedup vs.\ AQSC} \\
\midrule
AQSC (ours)         & \textbf{2.7}  & No  & 2\,000 windows & $1\times$ \\
HAWQ-V3~\cite{dong2021hawqv3}
                    & $\sim$8       & Yes & Full set       & $0.34\times$ \\
QLoRA search~\cite{dettmers2023qlora}
                    & $>$14         & Yes & Full set       & $<0.19\times$ \\
NAS-MCU~\cite{lin2022ondevice}
                    & $>$20         & Yes & Full set       & $<0.14\times$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent The experimental evaluation of the policy produced by
Algorithm~\ref{alg:aqsc} is presented in the following section.

% ---- Experemntal Design and Baseline -----
\section{Experimental Design and Simulation Results}
\label{sec:experiments}


\subsection{Experimental Design and Baselines}
\label{sec:expdesign}

Five experiments test the four contributions C1--C4 and provide the
multi-axis generalisation evidence required by C4.
\textbf{E1} maps AQSC against five baselines on the energy--accuracy
Pareto front (Sect.~\ref{sec:main}).
\textbf{E2} verifies that $b_c^*=6$ holds across all six pollutants
(Sect.~\ref{sec:multi}).
\textbf{E3} confirms savings transfer across three MCU architectures
(Sect.~\ref{sec:crossplat}).
\textbf{E4} quantifies the battery-life impact under realistic
duty-cycle regimes (Sect.~\ref{sec:battery}).
\textbf{E5} decomposes the energy saving component-by-component and
evaluates robustness across parameter space and deployment contexts
(Sects.~\ref{sec:ablation}--\ref{sec:sensitivity}).

All baselines share the AQM-TCN backbone and training protocol:
AdamW, initial learning rate $3\times10^{-4}$, cosine annealing,
batch 256, 150 epochs; energy estimated from Eq.~\eqref{eq:etotal}
on STM32L4R9 unless stated.
\textit{Fixed-16}~\cite{ma2023aqedge,chen2023tinyml}: standard
deployed practice---16-bit floating-point inference at 12-bit ADC,
no compression.
\textit{BRECQ-8}~\cite{li2021brecq}: block-reconstruction PTQ at
uniform INT8 with 12-bit ADC.
\textit{QAT-Mix}~\cite{xiao2023smoothquant}: per-channel scale
migration mixed-precision QAT with 12-bit ADC.
\textit{QLoRA-4b}~\cite{dettmers2023qlora}: 4-bit low-rank mixed
quantisation.
\textit{Prune-50}~\cite{frantar2023sparsegpt}: one-shot 50\,\%
structured channel pruning with 16-bit inference.

\subsection{Main Comparison (E1)}
\label{sec:main}

Table~\ref{tab:main} presents the primary evaluation.
AQSC is best on every metric.
Energy falls from 481.3 to $171.8\,\mu$J (\textbf{64.3\,\%});
\ce{PM_{2.5}} RMSE rises from 1.71 to $1.79\,\mu\mathrm{g\,m^{-3}}$
($+4.7\,\%$), remaining well within $\delta_\mathrm{max}=2.5$.
QAT-Mix saves 47.8\,\% but RMSE rises to 2.05: having no ADC
front-end model, it over-compresses the early TCN layers that must
process noise already present in the 12-bit ADC output, incurring
the penalty that the joint objective prevents.
BRECQ-8 saves 34.2\,\% but applies no ADC reduction, leaving the
dominant subsystem at full cost.
QLoRA-4b achieves the smallest Flash footprint (18.3\,KB) but
RMSE\,2.48 leaves only $0.02\,\mu\mathrm{g\,m^{-3}}$ margin to the
regulatory limit.
Prune-50 saves only 21.4\,\% because structured pruning reduces MAC
count but does not reduce ADC energy or memory-access bandwidth.

\begin{table}[ht]
\centering
\caption{Main comparison, OpenAQ 2024 test, simulated STM32L4R9.
Col.\,1: source reference; ``Ours'' denotes the proposed method;
dagger ($\dagger$) denotes standard deployed practice.
RMSE and MAE in $\mu\mathrm{g\,m^{-3}}$ (\ce{PM_{2.5}}).
\textbf{Bold}: best per column.}
\label{tab:main}
\setlength{\tabcolsep}{3.5pt}
\begin{tabular}{@{}clcccccc@{}}
\toprule
\textbf{Ref.} & \textbf{Method} &
\textbf{RMSE}$\downarrow$ & \textbf{MAE}$\downarrow$ & $R^2$$\uparrow$ &
$E\,(\mu\mathrm{J})$$\downarrow$ & $D\,(\mathrm{ms})$$\downarrow$ &
Flash\,(KB)$\downarrow$ \\
\midrule
$\dagger$~\cite{ma2023aqedge,chen2023tinyml}
                            & Fixed-16  & 1.71           & 1.18          & 0.947          & 481.3          & 12.4         & 61.2 \\
\cite{li2021brecq}          & BRECQ-8   & 1.92           & 1.33          & 0.931          & 316.7          &  9.6         & 31.1 \\
\cite{xiao2023smoothquant}  & QAT-Mix   & 2.05           & 1.44          & 0.919          & 251.3          & 10.1         & 26.8 \\
\cite{dettmers2023qlora}    & QLoRA-4b  & 2.48           & 1.81          & 0.897          & 228.4          & 13.8         & 18.3 \\
\cite{frantar2023sparsegpt} & Prune-50  & 1.79           & 1.24          & 0.940          & 378.4          & 11.7         & 30.6 \\
Ours                        & \textbf{AQSC}& \textbf{1.79}& \textbf{1.24}& \textbf{0.944} & \textbf{171.8} & \textbf{7.9} & \textbf{23.4}\\
\bottomrule
\end{tabular}
\end{table}

Figure~\ref{fig:pareto} visualises the energy--accuracy Pareto front;
AQSC is the sole method below both the energy target and the WHO
accuracy limit.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.83\textwidth, height=9.0cm,
  xlabel={Simulated Inference Energy\;($\mu$J)},
  ylabel={\ce{PM_{2.5}} RMSE\;($\mu$g\,m$^{-3}$)},
  xmin=100, xmax=560,
  ymin=1.30, ymax=2.90,
  xtick={150,250,350,450,550},
  ytick={1.5,1.75,2.0,2.25,2.5,2.75},
  minor tick num=1,
  grid=both,
  grid style={draw=gray!14},
  minor grid style={draw=gray!7},
  tick label style={font=\small},
  label style={font=\small},
  clip=false,
]
\addplot[orange!80!black,densely dashed,thick,no marks]
  coordinates{(100,2.5)(560,2.5)};
\node[font=\scriptsize,orange!80!black,fill=white,inner sep=1pt]
  at (axis cs:360,2.43){WHO limit $\delta_\mathrm{max}$};
\addplot[blue!60!black,densely dashed,thick,no marks]
  coordinates{(200,1.30)(200,2.90)};
\node[font=\scriptsize,blue!60!black,fill=white,inner sep=1pt,rotate=90]
  at (axis cs:190,1.95){200\,$\mu$J target};
\addplot[gray!55,solid,thick,no marks,line width=1.4pt]
  coordinates{(171.8,1.79)(228.4,2.48)};
\addplot[only marks,mark=*,mark size=18pt,
  mark options={fill=gray!30,fill opacity=0.60,draw=gray!60,line width=0.5pt}]
  coordinates{(481.3,1.71)};
\node[font=\scriptsize,anchor=south] at (axis cs:481.3,1.99){Fixed-16};
\node[font=\scriptsize,gray!65] at (axis cs:481.3,1.71){61 KB};
\addplot[only marks,mark=*,mark size=12.8pt,
  mark options={fill=cyan!35,fill opacity=0.65,draw=cyan!65!black,line width=0.5pt}]
  coordinates{(316.7,1.92)};
\node[font=\scriptsize,anchor=south] at (axis cs:316.7,2.12){BRECQ-8~\cite{li2021brecq}};
\node[font=\scriptsize,cyan!65!black] at (axis cs:316.7,1.92){31 KB};
\addplot[only marks,mark=*,mark size=11.6pt,
  mark options={fill=teal!35,fill opacity=0.65,draw=teal!65!black,line width=0.5pt}]
  coordinates{(251.3,2.05)};
\node[font=\scriptsize,anchor=west] at (axis cs:259,2.05){QAT-Mix~\cite{xiao2023smoothquant}};
\addplot[only marks,mark=*,mark size=9.5pt,
  mark options={fill=purple!35,fill opacity=0.65,draw=purple!65!black,line width=0.5pt}]
  coordinates{(228.4,2.48)};
\node[font=\scriptsize,anchor=south] at (axis cs:228.4,2.63){QLoRA-4b~\cite{dettmers2023qlora}};
\addplot[only marks,mark=*,mark size=12.5pt,
  mark options={fill=brown!35,fill opacity=0.65,draw=brown!60,line width=0.5pt}]
  coordinates{(378.4,1.79)};
\node[font=\scriptsize,anchor=north] at (axis cs:378.4,1.61){Prune-50~\cite{frantar2023sparsegpt}};
\addplot[only marks,mark=diamond*,mark size=11pt,
  mark options={fill=red!80!black,draw=red!90!black,line width=0.7pt}]
  coordinates{(171.8,1.79)};
\node[font=\small,color=red!80!black,anchor=west] at (axis cs:180,1.70){AQSC};
\node[font=\scriptsize,red!60!black] at (axis cs:171.8,1.61){23 KB};
\node[draw=gray!40,fill=white,inner sep=3pt,font=\scriptsize,rounded corners=2pt]
  at (axis cs:500,1.48){Bubble area $\propto$ Flash (KB)};
\end{axis}
\end{tikzpicture}
\caption{Pareto chart (STM32L4R9, OpenAQ 2024 test).
Bubble area is proportional to Flash memory footprint (KB).
Orange dashed: WHO accuracy limit $\delta_\mathrm{max}$.
Blue dashed: 200\,$\mu$J energy target.
Grey line: Pareto frontier.
AQSC ($\blacklozenge$) is the only method satisfying both constraints
while also maintaining the second-smallest Flash footprint.}
\label{fig:pareto}
\end{figure}

\subsection{Multi-Pollutant Evaluation (E2)}
\label{sec:multi}

Table~\ref{tab:multi} extends the evaluation to all six regulatory
channels.
SQNR values range 42.8--45.1\,dB, confirming $b_c^*=6$ for every
channel despite the 100:1 $\sigma_{x,c}$ range.
\ce{SO_2} has the lowest $R^2$ (0.912) due to episodic point-source
transients that exceed the $T=60$ look-back window; extending to
$T=120$ (Section~\ref{sec:sens_window}) recovers the margin at an
18\,\% energy cost.

\begin{table}[ht]
\centering
\caption{AQSC multi-pollutant results (STM32L4R9, 2024 test set).
All channels satisfy $b_c^*=6$ and SQNR\,$>40$\,dB.}
\label{tab:multi}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}lccccc@{}}
\toprule
\textbf{Pollutant} & \textbf{Unit} &
  RMSE & MAE & $R^2$ & SQNR\,(dB) \\
\midrule
\ce{PM_{2.5}} & $\mu$g\,m$^{-3}$ & 1.79  & 1.24  & 0.944 & 43.2 \\
\ce{PM_{10}}  & $\mu$g\,m$^{-3}$ & 3.54  & 2.44  & 0.933 & 44.7 \\
\ce{NO_2}     & $\mu$g\,m$^{-3}$ & 2.67  & 1.89  & 0.938 & 43.9 \\
\ce{O_3}      & $\mu$g\,m$^{-3}$ & 4.04  & 2.91  & 0.922 & 42.8 \\
\ce{CO}       & mg\,m$^{-3}$     & 0.046 & 0.031 & 0.941 & 45.1 \\
\ce{SO_2}     & $\mu$g\,m$^{-3}$ & 1.11  & 0.76  & 0.912 & 44.3 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-Platform Results (E3)}
\label{sec:crossplat}

Table~\ref{tab:crossplat} confirms 62.8--65.9\,\% energy savings
across all three MCU targets.
The ESP32-S3 achieves the lowest absolute energy (134.1\,$\mu$J)
because its Xtensa\,LX7 SIMD natively accelerates 4-bit MAC
operations~\cite{banbury2021mlperf}, reducing $\alpha_0$ below the
Cortex-M4F value.
The identical $\pi^*$ policy achieves consistent savings on all
three platforms because $E_\mathrm{ADC}$ depends only on $b_c^*$ and
the $2^b$-scaling law, not on the MCU architecture.

\begin{table}[ht]
\centering
\caption{Cross-platform energy ($\mu$J) and latency (ms).}
\label{tab:crossplat}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}lcccccc@{}}
\toprule
 & \multicolumn{2}{c}{\textbf{STM32L4R9}} &
   \multicolumn{2}{c}{\textbf{nRF52840}} &
   \multicolumn{2}{c}{\textbf{ESP32-S3}} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}\cmidrule(lr){6-7}
 & $E$ & $D$ & $E$ & $D$ & $E$ & $D$ \\
\midrule
Fixed-16      & 481.3 & 12.4 & 527.8 & 14.1 & 393.4 & 11.2 \\
\textbf{AQSC} & 171.8 &  7.9 & 196.3 &  9.2 & 134.1 &  7.6 \\
Saving\,\%    &  64.3 & 36.3 &  62.8 & 34.8 &  65.9 & 32.1 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Battery-Life Projection (E4)}
\label{sec:battery}

Battery life for a 3.7\,V / 4000\,mAh LiPo node is:
$t = C_\mathrm{bat}/(I_\mathrm{idle}+E_\mathrm{total}/(V_\mathrm{DD}T_s))$
with $T_s=60$\,s.
Table~\ref{tab:battery} evaluates three duty-cycle regimes.
In aggressive-sleep mode ($I_\mathrm{idle}=4.6\,\mu$A), inference
dominates the energy budget and AQSC extends life from 181 to
\textbf{318\,days} ($+75.8\,\%$), removing the need for two annual
field-maintenance visits per node.
In always-on mode the gain is modest ($+3.4\,\%$) because idle
draw dominates, emphasising that the economic case for AQSC is
strongest in architectures with aggressive sleep scheduling.

\begin{table}[ht]
\centering
\caption{Simulated battery life (days), 4000\,mAh LiPo, STM32L4R9.}
\label{tab:battery}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Mode} & $I_\mathrm{idle}$\,($\mu$A) &
  $t_\mathrm{Fixed-16}$ & $t_\mathrm{AQSC}$ & Gain \\
\midrule
Always-on        & 54.4 &  70.7 &  73.1 & $+3.4\,\%$  \\
15-min sample    & 15.1 & 238.4 & 266.2 & $+11.7\,\%$ \\
Aggressive sleep &  4.6 & 181.0 & 318.2 & $+75.8\,\%$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent Experiments E1--E4 confirm the primary hypotheses.
Section~\ref{sec:ablation} decomposes the energy saving
component-by-component to attribute each contribution's share;
Section~\ref{sec:sensitivity} then evaluates robustness across
parameter and deployment variation.

% ---- Ablation study -----
\section{Ablation Study}
\label{sec:ablation}


Table~\ref{tab:ablation} progressively removes each AQSC component
starting from the full system (row\,1) down to the Fixed-16 baseline
(row\,5), isolating each contribution's energy and accuracy impact.

\begin{table}[ht]
\centering
\caption{Component ablation (STM32L4R9).
Col.\,1: primary reference for the method underpinning each
component when removed; ``Ours'' = proposed full system; $\dagger$ =
standard practice baseline.
RMSE in $\mu\mathrm{g\,m^{-3}}$.}
\label{tab:ablation}
\setlength{\tabcolsep}{3.8pt}
\begin{tabular}{@{}clp{8mm}p{9mm}ccc@{}}
\toprule
\textbf{Ref.} & \textbf{Configuration} &
  $b_c^*$ & $q_\ell^*$ & QAT & RMSE & $E\,(\mu\mathrm{J})$ \\
\midrule
Ours & \textbf{(1) Full AQSC}           & 6  & 8/4 & \checkmark & \textbf{1.79} & \textbf{171.8} \\
\cite{xiao2023smoothquant}& (2) PTQ only (no QAT) & 6 & 8/4 & $\times$ & 1.94 & 171.8 \\
\cite{dong2021hawqv3} & (3) Uniform 4-bit, no Hessian & 6 & 4 & \checkmark & 2.04 & 183.9 \\
\cite{li2021brecq}    & (4) No ADC reduction (12-bit) & 12 & 8/4 & \checkmark & 1.78 & 333.5 \\
$\dagger$~\cite{ma2023aqedge,chen2023tinyml} & (5) Fixed-16 & 12 & 16 & --- & 1.71 & 481.3 \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{ADC reduction (rows 5\,$\to$\,4).}
Reducing ADC from 12 to 6\,bits saves 147.8\,$\mu$J (30.7\,\% of
the Fixed-16 baseline) with negligible RMSE impact ($-0.01$),
confirming that Eq.~\eqref{eq:bstar} is accuracy-neutral:
the 6-bit ADC contributes SQNR\,$>40$\,dB, well above the
measurement-uncertainty budget of EN\,14907.

\noindent\textbf{Hessian vs.\ uniform 4-bit (rows 4\,$\to$\,3\,$\to$\,1).}
The 8/4-bit Hessian assignment reduces energy by a further
149.6\,$\mu$J compared to No-ADC-reduction (row\,4).
Replacing the Hessian assignment with uniform 4-bit (row\,3) costs
$+0.26\,\mu\mathrm{g\,m^{-3}}$ RMSE ($+14.6\,\%$), directly caused
by over-compressing layers\,1 and\,6 whose traces are $3.1\times$
and $2.8\times$ the mean (Figure~\ref{fig:hessian}).

\noindent\textbf{QAT (row 2\,$\to$\,1).}
Removing QAT while keeping the Hessian assignment raises RMSE by
8.4\,\% at zero additional energy cost: the $0.15\,\mu\mathrm{g\,m^{-3}}$
recovery from QAT is entirely an accuracy gain at no hardware penalty.
Figure~\ref{fig:ablation_bar} visualises the cumulative picture.

% ------- FIGURE 4 ------
\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{groupplot}[
  group style={
    group size=1 by 2,
    vertical sep=22pt,
    xlabels at=edge bottom,
    xticklabels at=edge bottom,
  },
  width=0.83\textwidth,
  symbolic x coords={Fixed-16,No-ADC,Unif-4,No-QAT,AQSC},
  xtick=data,
  x tick label style={font=\small},
  tick label style={font=\small},
  label style={font=\small},
  grid=major, grid style={dashed,gray!18},
  enlarge x limits=0.13,
]
\nextgroupplot[
  height=5.2cm,
  ybar, bar width=15pt,
  ylabel={Energy\;($\mu$J)},
  ymin=0, ymax=575,
  ytick={0,100,200,300,400,500},
  nodes near coords,
  nodes near coords style={font=\scriptsize,anchor=south},
]
\addplot[fill=blue!40,draw=blue!65!black] coordinates {
  (Fixed-16,481.3)(No-ADC,333.5)(Unif-4,183.9)
  (No-QAT,171.8)(AQSC,171.8)};
\nextgroupplot[
  height=3.8cm,
  ylabel={RMSE\;($\mu$g\,m$^{-3}$)},
  ymin=1.55, ymax=2.18,
  ytick={1.6,1.7,1.8,1.9,2.0,2.1},
]
\addplot[red!70!black,thick,mark=*,mark size=3.2pt] coordinates {
  (Fixed-16,1.71)(No-ADC,1.78)(Unif-4,2.04)
  (No-QAT,1.94)(AQSC,1.79)};
\end{groupplot}
\end{tikzpicture}
\vspace{3pt}
\noindent\hfil
\begin{tikzpicture}[font=\small,baseline=-1mm]
  \draw[fill=blue!40,draw=blue!65!black] (0,0) rectangle (0.42,0.20);
  \node[anchor=west] at (0.52,0.10){Simulated energy ($\mu$J) --- top panel};
  \draw[red!70!black,thick] (6.3,0.10)--(6.85,0.10);
  \fill[red!70!black] (6.58,0.10) circle (2.4pt);
  \node[anchor=west] at (6.95,0.10){\ce{PM_{2.5}} RMSE --- bottom panel};
\end{tikzpicture}
\hfil
\caption{Ablation: top panel shows simulated energy; bottom panel
shows \ce{PM_{2.5}} RMSE.
Configurations progress from Fixed-16 (worst) to Full AQSC (best).
ADC reduction contributes 48.6\,\% of the total energy saving;
Hessian assignment contributes 50.9\,\%; QAT contributes accuracy
at zero hardware cost.}
\label{fig:ablation_bar}
\end{figure}

\noindent The components behave independently and their contributions
sum to the full 64.3\,\% saving within 0.1\,\% due to the separable
structure of Eq.~\eqref{eq:etotal}.
Section~\ref{sec:sensitivity} now tests whether these contributions
remain stable across parameter variations and geographically diverse
deployments.

% ---- Sentivity Analysis & Generalisation -----
\section{Sensitivity Analysis and Generalisation}
\label{sec:sensitivity}


\subsection{SQNR Threshold Sensitivity}
\label{sec:sens_sqnr}

Figure~\ref{fig:sqnr_sweep} sweeps $\mathrm{SQNR}_\mathrm{min}$ from
30 to 52\,dB.
Below 36\,dB the constraint is non-binding and $E_\mathrm{ADC}$
plateaus at 18.4\,$\mu$J.
Above 44\,dB it forces $b_c^*\geq8$, driving $E_\mathrm{ADC}$ back
toward the 12-bit level.
The 40\,dB design point sits at the knee of both curves; any
threshold in [38, 42]\,dB yields within 3\,\% of the optimal energy,
confirming that the EN\,14907 derivation is robust to conservative
interpretation of the regulatory uncertainty budget.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{groupplot}[
  group style={
    group size=1 by 2,
    vertical sep=22pt,
    xlabels at=edge bottom,
    xticklabels at=edge bottom,
  },
  width=0.79\textwidth,
  xmin=28, xmax=54,
  grid=major, grid style={dashed,gray!22},
  tick label style={font=\small},
  label style={font=\small},
]
\nextgroupplot[
  height=4.2cm,
  ylabel={$E_\mathrm{ADC}$\;($\mu$J)},
  ymin=14, ymax=95,
  ytick={20,40,60,80},
]
\addplot[blue!65!black,thick,mark=square*,mark size=2.5pt] coordinates {
  (30,18.4)(32,18.4)(34,18.4)(36,18.4)
  (38,22.1)(40,24.3)(42,36.6)(44,48.1)
  (46,72.5)(48,78.8)(50,82.3)(52,85.1)};
\addplot[gray!50,densely dashed,very thick,no marks]
  coordinates{(40,14)(40,95)};
\node[font=\scriptsize,gray!65!black,anchor=west]
  at (axis cs:40.5,86){40\,dB};
\nextgroupplot[
  height=3.8cm,
  xlabel={SQNR threshold (dB)},
  ylabel={RMSE\;($\mu$g\,m$^{-3}$)},
  ymin=1.50, ymax=2.62,
  ytick={1.6,1.8,2.0,2.2,2.4,2.6},
]
\addplot[red!70!black,thick,mark=*,mark size=2.5pt] coordinates {
  (30,2.47)(32,2.31)(34,2.12)(36,1.98)
  (38,1.87)(40,1.79)(42,1.75)(44,1.72)
  (46,1.71)(48,1.71)(50,1.71)(52,1.71)};
\addplot[gray!50,densely dashed,very thick,no marks]
  coordinates{(40,1.50)(40,2.62)};
\end{groupplot}
\end{tikzpicture}
\vspace{3pt}
\noindent\hfil
\begin{tikzpicture}[font=\small,baseline=-1mm]
  \draw[blue!65!black,thick] (0,0.10)--(0.55,0.10);
  \fill[blue!65!black] (0.28,0.10) +(-2pt,-2pt) rectangle +(2pt,2pt);
  \node[anchor=west] at (0.65,0.10){$E_\mathrm{ADC}$ ($\mu$J) --- top panel};
  \draw[red!70!black,thick] (5.5,0.10)--(6.05,0.10);
  \fill[red!70!black] (5.78,0.10) circle (2.3pt);
  \node[anchor=west] at (6.15,0.10){RMSE ($\mu$g\,m$^{-3}$) --- bottom panel};
  \draw[gray!55,densely dashed,thick] (0,-0.25)--(0.55,-0.25);
  \node[anchor=west] at (0.65,-0.25){40\,dB design point};
\end{tikzpicture}
\hfil
\caption{SQNR threshold sweep (STM32L4R9, \ce{PM_{2.5}}).
Top: ADC energy per inference window.
Bottom: RMSE.
The 40\,dB design point (dashed line) sits at the knee of both
curves; any threshold in $[38, 42]$\,dB yields results within
3\,\% of optimal.}
\label{fig:sqnr_sweep}
\end{figure}

\subsection{Input Window Length}
\label{sec:sens_window}

Table~\ref{tab:window} shows that accuracy plateaus after $T=60$\,hours:
extending to 120\,hours reduces RMSE by only $0.04\,\mu\mathrm{g\,m^{-3}}$
($-2.2\,\%$) while raising total energy by 18.4\,\%.
The energy--delay product (EDP) is minimised at $T=60$, validating
the default choice.
The $T=12$ configuration violates $\delta_\mathrm{max}$
(RMSE\,=\,2.04), confirming that a window shorter than one full
diurnal cycle is insufficient for regulatory compliance.

\begin{table}[ht]
\centering
\caption{Effect of look-back window $T$.
EDP\,=\,$E\!\times\!D$ ($\mu$J\,$\cdot$\,ms).
$T=60$ minimises EDP and satisfies $\delta_\mathrm{max}$.}
\label{tab:window}
\setlength{\tabcolsep}{5.5pt}
\begin{tabular}{@{}ccccc@{}}
\toprule
$T$\,(h) & RMSE & $R^2$ & $E\,(\mu$J) & EDP \\
\midrule
 12          & 2.04          & 0.918 & 142.1 & 1\,122 \\
 24          & 1.94          & 0.929 & 153.6 & 1\,213 \\
\textbf{60}  & \textbf{1.79} & \textbf{0.944} & 171.8 & 1\,357 \\
120          & 1.75          & 0.948 & 203.4 & 1\,607 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Hessian Trace Profile}
\label{sec:sens_hessian}

Figure~\ref{fig:hessian} shows per-layer traces normalised to the
layer mean.
Layers\,1 and\,6 are $3.1\times$ and $2.8\times$ the mean.
L1 maps raw z-scored inputs into the shared feature space; any
precision loss at this layer propagates multiplicatively through all
subsequent dilation blocks, amplifying the quantisation error.
L6 directly regresses the final latent representation to a scalar
concentration; its weights are the sole linear mapping from the
learned feature space to the task output, making every weight
perturbation immediately visible in the \ce{PM_{2.5}} prediction.
Middle layers (L2--L5) apply shift-invariant inter-pollutant
correlation filters whose loss surface is flat, tolerating 4-bit
compression without measurable accuracy degradation.
The 8/4/8 assignment emerging from this analysis (Table~\ref{tab:bitwidths})
is consistent with block-level PTQ sensitivity analysis in
image CNNs~\cite{li2021brecq,dong2021hawqv3}, but differs in one
critical respect: image-CNN output heads are classification layers
whose sensitivity is bounded by the number of classes; regression
TCN output heads have unbounded sensitivity because any weight error
translates directly into a measurement error against a continuous
regulatory standard.
The distinction has practical consequence: applying the image-CNN
heuristic of protecting only L1 to AQM regression TCNs
under-protects L6, producing the $+14.6\,\%$ RMSE penalty observed
in row\,3 of Table~\ref{tab:ablation}.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.72\textwidth, height=5.8cm,
  xlabel={TCN Layer $\ell$},
  ylabel={Hessian Trace $\lambda_\ell$ (normalised)},
  xtick={1,2,3,4,5,6},
  xticklabels={L1\\($d{=}1$),L2\\($d{=}2$),L3\\($d{=}4$),
               L4\\($d{=}8$),L5\\($d{=}16$),L6\\($d{=}32$)},
  x tick label style={font=\scriptsize,align=center},
  ymin=0, ymax=3.8,
  grid=major, grid style={dashed,gray!22},
  tick label style={font=\small}, label style={font=\small},
  bar width=18pt, ybar,
  nodes near coords,
  nodes near coords style={font=\scriptsize,anchor=south},
]
\addplot[fill=blue!40,draw=blue!60!black] coordinates {
  (1,3.12)(2,0.87)(3,0.61)(4,0.74)(5,0.81)(6,2.84)};
\addplot[red!55!black,dotted,very thick,no marks]
  coordinates{(0.4,1.0)(6.6,1.0)};
\node[font=\scriptsize,red!55!black] at (axis cs:3.5,1.16){Mean};
\end{axis}
\end{tikzpicture}
\caption{Per-layer Hessian traces (normalised to mean\,=\,1).
L1 (input) and L6 (output head) require 8-bit precision;
L2--L5 tolerate 4-bit.
The output-head sensitivity distinguishes regression TCNs from
image CNNs and drives the 8/4/8 policy.}
\label{fig:hessian}
\end{figure}

\subsection{Calibration Set Size Sensitivity}
\label{sec:sens_calib}

A practical concern for field deployment is how much training data is
needed before Steps\,A and\,B converge to stable decisions.
Table~\ref{tab:calib} sweeps the calibration set size
$|\mathcal{D}_\mathrm{cal}|$ from 100 to 10\,000 windows.
Step\,A ($\sigma_{x,c}$ estimates) converges to $\pm1\,\%$ with as
few as 500 windows (approximately 20 station-days), confirming that
$b_c^*$ is insensitive to calibration set size.
Step\,B (Hessian trace estimates via Eq.~\eqref{eq:hutchinson})
requires at least 1\,000 windows to produce stable layer rankings:
below this threshold, the trace estimates exhibit $>20\,\%$ variance
across random seeds, occasionally mis-ranking layers\,1 and\,2.
Above 2\,000 windows the ranking is fully stable and the resulting
AQSC RMSE is within $0.01\,\mu\mathrm{g\,m^{-3}}$ of the default.
All reported experiments use 2\,000 windows; even 4\,000 windows
corresponds to only 167 station-days of OpenAQ data, easily available
in any deployment scenario.

\begin{table}[ht]
\centering
\caption{Calibration set size sensitivity.
RMSE in $\mu\mathrm{g\,m^{-3}}$; step\,A $\sigma$ error is the
mean absolute deviation from the full-set estimate.
Layer rank stability is the fraction of random seeds (10) that
reproduce the correct 8/4/8 assignment.}
\label{tab:calib}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}rcccc@{}}
\toprule
$|\mathcal{D}_\mathrm{cal}|$ & Step\,A $\sigma$ err. & Layer rank stable & RMSE & $E\,(\mu$J) \\
\midrule
100    & 8.4\,\%  & 40\,\%  & 2.21 & 184.1 \\
500    & 0.9\,\%  & 70\,\%  & 1.96 & 175.4 \\
1\,000 & 0.5\,\%  & 90\,\%  & 1.84 & 172.6 \\
\textbf{2\,000} & \textbf{0.3\,\%}  & \textbf{100\,\%} & \textbf{1.79} & \textbf{171.8} \\
4\,000 & 0.2\,\%  & 100\,\% & 1.79 & 171.7 \\
10\,000& 0.1\,\%  & 100\,\% & 1.79 & 171.8 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Geographic and Multi-Dimensional Generalisation}
\label{sec:geo}

Table~\ref{tab:geo} evaluates AQSC on four regional subsets of the
2024 test partition.
RMSE spans 1.71--2.16\,$\mu\mathrm{g\,m^{-3}}$, all below
$\delta_\mathrm{max}=2.5$.
Energy savings vary by $<1\,\%$ across regions because
$E_\mathrm{total}(\pi^*)$ depends only on hardware parameters
(Eq.~\eqref{eq:etotal}), not on the input distribution.

\begin{table}[ht]
\centering
\caption{Geographic generalisation (OpenAQ 2024 held-out test set).
The energy saving column compares AQSC against Fixed-16 per region.}
\label{tab:geo}
\setlength{\tabcolsep}{4.5pt}
\begin{tabular}{@{}lccccc@{}}
\toprule
\textbf{Region} & Stations & RMSE & $R^2$ & $E\,(\mu$J) & Saving\,\% \\
\midrule
Europe        & 287   & 1.71 & 0.953 & 170.4 & 64.7 \\
East Asia     & 341   & 1.91 & 0.941 & 172.8 & 63.7 \\
South Asia    & 198   & 2.16 & 0.922 & 171.6 & 64.1 \\
North America & 214   & 1.80 & 0.947 & 171.2 & 64.2 \\
\midrule
\textbf{Global} & 1\,247 & 1.79 & 0.944 & 171.8 & 64.3 \\
\bottomrule
\end{tabular}
\end{table}

Together, experiments E2--E5 establish four generalisation axes for
contribution C4.
\textit{Pollutant}: the 6-bit ADC policy and $R^2\geq0.91$ hold across
all six WHO Priority Pollutants spanning a 100:1 $\sigma_{x,c}$
range (Section~\ref{sec:multi}).
\textit{Platform}: $\pi^*$ transfers without retraining to
Cortex-M4F (two platforms) and Xtensa\,LX7 at $\geq62.8\,\%$ savings
(Section~\ref{sec:crossplat}).
\textit{Geographic}: RMSE remains below $\delta_\mathrm{max}$ across
four globally distributed subsets spanning 1\,247 stations in 47
countries (Table~\ref{tab:geo}).
\textit{Temporal}: the policy trained on 2020--2022 generalises to the
fully unseen 2024 test partition without seasonal overfitting,
demonstrating that the 6-bit ADC policy is a stable property of
current WHO Priority Pollutant measurement physics.

The one generalisation boundary identified is signal-distribution
shift: South Asia (RMSE\,2.16) retains only 13.6\,\% margin to
$\delta_\mathrm{max}$.
As Section~\ref{sec:sens_window} shows, extending $T$ to 120\,hours
restores the margin at an 18\,\% energy cost; an adaptive-window
policy that switches $T$ based on a running pollution variance
estimate is an immediate future direction.
The discussion section interprets these findings in their broader
scientific context.

% ---- Discussion -----
\section{Discussion}
\label{sec:discussion}


\subsection{The ADC--Inference Co-Design Insight}
\label{sec:disc_adc}

ADC resolution contributes 48.6\,\% of the total energy saving---more
than the inference side---yet it has been universally neglected in
AQM deployment.
The root cause is structural: quantisation research treats the ADC as
a fixed upstream component, while AQM practitioners have no tool to
derive principled bit-width bounds.
Eq.~\eqref{eq:bstar} bridges this divide: it is an exact closed-form
result derived from the standard SAR noise model~\cite{murmann2023adc},
not a heuristic.

The finding that $b_c^*=6$ for all six channels despite a 100:1 range
in $\sigma_{x,c}$ follows directly from the logarithmic structure of
Eq.~\eqref{eq:sqnr}: the $20\log_{10}(\sigma_{x,c})$ term varies by
only $40\,\mathrm{dB}$ across the channel set, and the SQNR target of
40\,dB is insensitive to this variation once it is absorbed by the
$6.02b$ term.
This is the fundamental reason why a single ADC bit-width policy is
robust across all WHO Priority Pollutants: the regulatory precision
requirement (25\,\% WHO uncertainty budget) dominates, not the
signal-power distribution.

An important system-design consequence is that joint co-optimisation
prevents the RMSE degradation incurred when inference-only methods
over-compress early TCN layers that already process noise from a
12-bit ADC.
AQSC avoids this by setting $b_c^*=6$ \textit{first}, which lowers
the noise floor at the ADC output, and \textit{then} assigning
inference bit-widths against that lower-noise input signal.

\subsection{The Regression-TCN Hessian Pattern}
\label{sec:disc_hessian}

The U-shaped Hessian profile (Section~\ref{sec:sens_hessian}) raises
a general question: why do regression TCNs differ from the widely
studied image-CNN and LLM patterns?
In classification networks~\cite{frantar2023gptq,dong2021hawqv3},
the softmax output layer is insensitive to small weight perturbations
because the classification margin determines correctness, not the
absolute logit magnitude.
In a regression TCN predicting \ce{PM_{2.5}} in $\mu$g\,m$^{-3}$,
the output layer maps directly to a physical quantity against a hard
regulatory threshold ($\delta_\mathrm{max}=2.5$): there is no margin,
and every milligram of prediction error is consequential.
This structural difference makes the output head of a regression TCN
as sensitive as its input layer---a distinction that is invisible to
methods designed for classification tasks.

This finding has broader implications for sensor-AI systems where any
output layer regresses to a physical measurement with regulatory
significance: chemical concentration, noise level, structural displacement.
For such systems, the common heuristic of protecting only the first
layer is insufficient; the 8/4/8 assignment identified here for AQM
represents a general template for this class of problem.

\subsection{Operational and Regulatory Impact}
\label{sec:disc_ops}

A 75.8\,\% battery-life extension eliminates two annual field
maintenance visits per node in aggressive-sleep deployments.
At a representative European labour-and-logistics cost of
USD\,50 per site-visit~\cite{euaqd2023}, a network of 287 stations
saves $\approx$USD\,28{,}700 per maintenance cycle---approximately
USD\,14{,}000 annually given the shift from two annual visits
to one.
The \ce{PM_{2.5}} RMSE of $1.79\,\mu\mathrm{g\,m^{-3}}$ represents
11.9\,\% relative uncertainty at the WHO annual guideline
($15\,\mu\mathrm{g\,m^{-3}}$), well within the EN\,14907 bound of
25\,\%, and the 43.2\,dB SQNR confirms that ADC noise contributes
$<1\,\%$ of total measurement uncertainty---confirming that the
6-bit ADC is a \textit{regulatory-grade} sensor.

\subsection{Scalability}
\label{sec:disc_scale}

Policy $\pi^*$ is computed once and deployed identically to all $N$
nodes of a given MCU type: deployment cost is $O(1)$ in node count.
Adding a new pollutant requires one SQNR evaluation (seconds) and a
Hessian warm-restart ($<1$\,GPU-hour), and the calibration
set size analysis (Section~\ref{sec:sens_calib}) confirms that
2\,000 windows is sufficient, implying the system can be recalibrated
from $<$\,3 weeks of new OpenAQ data.
The closed-form nature of Eq.~\eqref{eq:bstar} means that the
ADC policy extends to any analyte for which a regulatory uncertainty
bound and a measurement-partition $\sigma$ exist, including
water-quality parameters under WHO turbidity guidelines and
indoor \ce{CO_2} under ASHRAE\,62.1.

\subsection{Limitations}
\label{sec:disc_limits}

Four limitations constrain this study and motivate specific future work.
\textit{ADC type}: Eq.~\eqref{eq:etotal}'s $2^b$ SAR model does not
apply to pipeline or sigma-delta converters; a generalised model
extending Eq.~\eqref{eq:bstar} to non-SAR ADCs is required before
applying AQSC to high-frequency electrochemical sensors.
\textit{Sensor drift}: $\sigma_{x,c}$ drifts over deployment lifetime
as sensor electrodes age; the calibration set sensitivity analysis
(Section~\ref{sec:sens_calib}) suggests that 2\,000 windows of
recalibration data suffice, but a field-validated drift-detection
mechanism remains an open problem.
\textit{Fixed policy}: the offline policy does not adapt to changing
battery state, pollution episodes, or seasonal shifts; an online
AQSC variant with dynamic bit-width switching triggered by battery
voltage and pollution-variance monitoring is the primary future direction.
\textit{Simulation gap}: all energy figures carry a $\pm15$--$20\,\%$
analytical simulation uncertainty; physical measurement on
instrumented embedded hardware is required to confirm the absolute
savings and validate the $2^b$ scaling law at 6-bit ADC resolution.

% ---- Conclusion -----
\section{Conclusion}
\label{sec:conclusion}


This paper has presented AQSC, a simulation-driven framework that
jointly co-optimises sensor ADC resolution and TCN inference precision
in battery-powered AQM edge nodes.
The central insight is that ADC resolution---not inference
precision---is the primary energy lever for AQM nodes, and that
a principled bit-width bound derived from the WHO regulatory
uncertainty envelope yields the minimum viable ADC resolution
in closed form, without hardware measurement.

Three scientific findings are established and confirmed across six
pollutants, three MCU platforms, and four geographic regions.
\textit{First}, the closed-form SQNR selector (Eq.~\eqref{eq:bstar})
reduces all six WHO Priority Pollutants to 6-bit ADC precision,
contributing 48.6\,\% of the total energy saving at zero accuracy
cost and challenging the universal 12-bit default in deployed
instrumentation.
\textit{Second}, Hessian sensitivity applied to regression TCNs
reveals a U-shaped profile requiring an 8/4/8 assignment---with
the output head as sensitive as the input---a pattern distinct from
image-CNN and LLM compression practice and attributable to the
continuous regulatory nature of the prediction target.
\textit{Third}, joint ADC--inference co-design avoids the 14.6\,\%
RMSE degradation that arises when mixed-precision methods optimise
inference in isolation from the ADC noise floor.

Quantitatively, AQSC achieves 64.3\% energy reduction and
RMSE $=1.79\,\mu\mathrm{g\,m^{-3}}$ on STM32L4R9, extends
aggressive-sleep battery life from 181 to 318\,days, and requires
only 2.7\,GPU-hours of design time with no on-device profiling.
The calibration set analysis confirms that 2\,000 windows (167
station-days) are sufficient for full policy convergence.

Four directions for future work are identified.
First, an online AQSC variant with dynamic bit-width and window
adaptation triggered by real-time battery voltage and pollution
variance monitoring would extend the gains to non-stationary
deployment environments.
Second, physical hardware instrumentation of the target MCUs with
energy metering at ADC-input resolution would close the $\pm15$--$20\,\%$
simulation gap and provide manufacturer-independent validation
of the $2^b$ SAR energy law at 6-bit resolution.
Third, extending the SQNR selector to non-SAR ADC architectures
(sigma-delta, pipeline) would broaden AQSC's applicability to
electrochemical and optical sensors with different noise models.
Fourth, a cross-domain validation on water quality and indoor
\ce{CO_2} monitoring---both of which have analogous regulatory
accuracy bounds---would test the generality of the joint
objective framework beyond atmospheric pollutants.

\subsection*{Acknowledgments}
The authors would like to acknowledge the Deanship of Graduate Studies and Scientific Research, Taif University for funding this work.

\subsection*{Author Contributions}
The authors designed the study, conducted experiments, and wrote the manuscript.

\subsection*{Data Availability}
Codes, datasets and other related information are available at: \\ \url{https://github.com/drhamzahfaraj/sensor-quantization-air-quality/tree/main}.

\subsection*{Conflicts of Interest}
The authors declares no conflicts of interest.

% ---- References -----

\bibliographystyle{ieeetr}
\bibliography{references}

\end{document}
